---
title: "Access geospatial data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Access geospatial data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE}
library(aemetools)
library(sf) # Used for spatial data
library(terra) # Used for raster data
library(tmap) # Used for spatial plotting

# tmap settings
tmap_mode("view")
tmap_options(basemap.server = "Esri.WorldImagery", raster.max_cells = 1e8)

```

This vignette demonstrates how to access geospatial data from various sources
including the [Land Information New Zealand (LINZ) Data Service](https://data.linz.govt.nz/)
and 
[Stats NZ Tatauranga Aotearoa (New Zealand's official data agency)](https://datafinder.stats.govt.nz/).

# Access shapefile data from Stats NZ Tatauranga Aotearoa

The Stats NZ Tatauranga Aotearoa website provides access to a range of geospatial
data including shapefiles. The `read_web_sf` function can be used to access this
data.

Here we will access the [regional boundaries shape file](https://datafinder.stats.govt.nz/layer/111182-regional-council-2023-generalised/)
from Stats NZ.

```{r read-stats-nz-shapefile, warning=FALSE, message=FALSE}
# Read the shapefile data from Stats NZ
url <- "https://datafinder.stats.govt.nz/" # Base URL
layer_id <- 111182 # Layer ID for the regional shapefile
key <- Sys.getenv("STATS_NZ_KEY") # Stats NZ API key

# Read the shapefile data
nz_shapefile <- read_web_sf(url = url, layer_id = layer_id, key = key)
nz_shapefile
```

### Visualise the shapefile data

This can be easily visualised using the `tmap` package.

These regions are the boundaries of the regional councils in New Zealand. 
However the Chatham islands are across the international date line and can make
plotting difficult. We will remove these from the shapefile.

```{r remove-chatham-islands}
# Remove the Chatham Islands from the shapefile
nz_shapefile <- nz_shapefile |> 
  dplyr::filter(C2023_V != "Area Outside Region")
```


```{r view-nz-shapefile}
# View the shapefile data
tm_shape(nz_shapefile, name = "Regional Council boundaries") +
  tm_borders(col = "blue") +
  tm_fill(fill = "C2023_V", 
          fill.legend = tm_legend(title = "Region"),
          fill_alpha = 0.3, id = "C2023_V")

```

# Access elevation data from LINZ

We will use Lake Rotoroa as an example to demonstrate how to access elevation 
data from the LINZ Data Service.

## Define the location of the lake

We will define the location of Lake Rotoroa and create a spatial dataframe of
a point in the lake using the `sf` package.

```{r lake-location}
# Define the location of the lake
lat <- -37.79881
lon <- 175.2742
lake <- "Lake Rotoroa"


# Create a spatial dataframe of the lake point
pnt <- sf::st_point(c(lon, lat)) |> 
  sf::st_sfc(crs = 4326) |> 
  sf::st_sf() |> 
  sf::st_set_geometry(value = "geometry") |> 
  dplyr::mutate(title = lake)
pnt

```

### Visualise the location of the lake

We can visualise the location of the lake using the [`tmap`](https://r-tmap.github.io/tmap/) package.

```{r view-lake-location}
# View the location of the lake in a map
map <- tm_shape(pnt, name = "Lake") +
  tm_dots(col = "blue", size = 0.2, popup.vars = "title") 

map +
  tm_view(set_view = 15)
```

## Query DEM metadata for elevation

### DEM metadata

Within the aemetools package, there is a spatial features (sf) dataframe which 
contains metadata about various Digital Elevation Models (DEMs) for New Zealand 
which are available on the LINZ Data Service. We can use this to find the
elevation of Lake Rotoroa.

```{r view-dem-metadata, eval = FALSE}
nz_dem_metadata
```

```{r view-dem-metadata-DT, echo = FALSE}

df <- nz_dem_metadata 
spl <- "\\([^()]*\\)(*SKIP)(*F)|\\h*\\.\\h*"
df$summary <- apply(df, 1, function(x) strsplit(x[["abstract"]], spl,
                                                perl = TRUE)[[1]][1])

df |> 
  sf::st_drop_geometry() |>
  dplyr::select(layer_id, title, summary, region, res, units, year) |>
  DT::datatable(options = list(pageLength = 4, scrollX = TRUE))
```

### View DEM coverage

We can view the coverage of the DEM layers to see which layers intersect with
our lake.

```{r view-dem-coverage, warning = FALSE}
# View the coverage of the DEM layers
tm_shape(nz_dem_metadata, name = "DEM coverage") +
  tm_polygons(fill = "red", col = "black", fill_alpha = 0.1, id = "title") +
  tm_shape(pnt, name = "Lake") +
  tm_dots(col = "blue", size = 0.2, popup.vars = "title") 

```

First we will transform our lake coordinates to NZTM and use this to find which 
layers intersect with  our lake from the DEM metadata using the 
`sf::st_intersects` function.

```{r subset-dem-layers}
# Subset the DEM database for the elevation of the lake  
pnt_nz <- pnt |>
  sf::st_transform(crs = 2193)

# Subset the DEM layers that intersect with the lake
dem_layers <- nz_dem_metadata[sf::st_intersects(pnt_nz, nz_dem_metadata, 
                                                sparse = FALSE), ]
dem_layers[, c("title", "region")]
```

There are two layers that intersect with the lake. We will do a quick visual
inspection of the layers to see which one is the most suitable for our
lake.

```{r view-dem-tmap}
# View the layers
dem_84 <- dem_layers |> 
  sf::st_transform(crs = 4326) 

tm_shape(dem_84) +
  tm_polygons(fill = "red", col = "black", popup.vars = "title", fill_alpha = 0.3) +
  # tm_borders(col = "red", popup.vars = "title") +
  tm_shape(pnt, name = "Lake") +
  tm_dots(col = "blue", size = 0.2, popup.vars = "title") +
  tm_view(set_view = c(lon, lat, 7))

```

## Querying the elevation values

For raster data, point queries (lat/lon) can be sent to the LINZ API. We can 
query the elevation values for the lake from the different layers using the 
`get_raster_layer_value` function. This function takes the latitude and
longitude of the point, and the layer id of the raster layer. We will compare
the elevation values from the different layers.

```{r query-elevation-values}
# Compare the lake elevation values from the different layers
val1 <- get_raster_layer_value(lat = lat, lon = lon, 
                               layer = dem_layers$layer_id[1])

val2 <- get_raster_layer_value(lat = lat, lon = lon, 
                               layer = dem_layers$layer_id[2])

data.frame(layer = dem_layers$title[1:2], elevation = c(val1, val2))

```

There is a discrepancy between the elevation values from the two layers. We will
use the layer from the second layer because it is a higher resolution (1m) and 
is from more recent data (2019).

<!-- ### Download the DEM raster data

We can download the LiDAR raster data for this layer and plot this over the lake
to see which one works

```{r plot-dem, warning = FALSE, message = FALSE}
dem <- get_raster_tile(x = pnt, layer_id = dem_layers$layer_id[2],
                       zoom = 18) |> 
  terra::subset(1) # All colours are the same, so we will subset the first layer
names(dem) <- "Elevation"

tm_shape(dem, raster.downsample = FALSE, name = "LiDAR") +
  tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE, 
            title = "Elevation (m)") 
```
-->

## Aerial imagery

Aerial imagery is also available from the LINZ Data Service. We can use this to
view the lake and its surroundings.

Within the `aemetools` package, there is also spatial features (sf) dataframe 
which contains metadata about various aerial imagery for New Zealand 
which are available on the LINZ Data Service. We can use this to source an 
aerial image of the lake.

```{r query-aerial}
# Query the aerial imagery metadata for the elevation of the lake  
# Subset the aerial imagery layers that intersect with the lake
aerial_layers <- nz_aerial_imagery_metadata[sf::st_intersects(pnt_nz, nz_aerial_imagery_metadata, 
                                                sparse = FALSE), ]
aerial_layers
```

```{r view-aerial-metadata, eval = FALSE}
aerial_layers
```

Here we can view the metadata of the aerial imagery layers that intersect with
our lake.

```{r view-aerial-metadata-DT, echo = FALSE}

df <- aerial_layers 
spl <- "\\([^()]*\\)(*SKIP)(*F)|\\h*\\.\\h*"
df$summary <- apply(df, 1, function(x) strsplit(x[["abstract"]], spl,
                                                perl = TRUE)[[1]][1])

df |> 
  sf::st_drop_geometry() |>
  dplyr::select(layer_id, title, summary, region, res, units, year) |>
  DT::datatable(options = list(pageLength = 5, scrollX = TRUE))

```

We can view the coverage of the aerial imagery layers to see which layers
intersect with our lake.

```{r view-aerial-coverage, warning = FALSE}
# View the coverage of the aerial imagery layers
tm_shape(aerial_layers, name = "Aerial imagery coverage") +
  tm_polygons(fill = "yellow", col = "black", fill_alpha = 0.1, id = "title") +
  tm_shape(pnt, name = "Lake") +
  tm_dots(col = "blue", size = 0.2, popup.vars = "title") 
```

We can download two aerial images for the lake and compare them.

It is important to note that the aerial imagery resolution is dependent on the
zoom level. The higher the zoom level, the higher the resolution. We will use
the level 17, as the higher the resolution, the longer it takes to download the
image.

This is important to note as even though some layers may have a higher
resolution, we may not need that resolution for our analysis. We can compare the


```{r plot-aerial-image, warning = FALSE, message = FALSE}
aerial1 <- get_raster_tile(x = pnt, layer_id = aerial_layers$layer_id[1],
                           zoom = 16) 
aerial2 <- get_raster_tile(x = pnt, layer_id = aerial_layers$layer_id[4],
                           zoom = 16) 

tm_shape(aerial1) +
  tm_rgb() +
  tm_shape(aerial2) +
  tm_rgb()
```

### Compare aerial imagery resolution

We can compare the resolution of the two aerial images to see which one is more
suitable for our lake.

```{r compare-aerial-resolution}
# Compare the resolution of the aerial imagery but first transform to NZTM
res1 <- aerial1 |> 
  terra::project("+proj=nzmg") |> 
  terra::res()

res2 <- aerial2 |> 
  terra::project("+proj=nzmg") |> 
  terra::res()

data.frame(layer = aerial_layers$title[c(1, 4)],
           resolution = c(res1[1], res2[1]))

```

## Access LINZ Aerial Basemap

We can also download the [LINZ Aerial Basemap](https://basemaps.linz.govt.nz/) 
to see the coverage of the lake. You will need to sign up for a free API key to
use this service.

```{r linz-basemap, warning = FALSE, message = FALSE}
# Get the LINZ basemap
linz_basemap <- get_linz_basemap_tile(x = pnt, zoom = 16)

tm_shape(linz_basemap, name = "LINZ Aerial Basemap") +
  tm_rgb()
```

