---
title: "Calibration strategies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibration strategies}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
unlink("calib-test", recursive = TRUE, force = TRUE)
ggplot2::update_geom_defaults("point", list(size = 0.3))
```

## Calibration strategies for Aquatic Ecosystem Models

Calibrating aquatic ecosystem models is a difficult task. The models are
complex, and the data are often sparse and noisy.

Optimising a model is a process of finding the best set of parameters. This can
potentiall be across a wide range of response variables. For example, it can
include optimisations for water temperature, lake level, and chlorophyll-a.
Therefore, it is important to have a strategy for calibrating the models under
such scenarions.

This vignette will demonstrate how to calibrate the AEME models using different
strategies.

### Strategy #1: Calibrate lake level

When running long-term simulations for lakes, it is critically important that
the lake water balance is fully resolved. This is because the lake level will be
a key driver for many of the processes in the lake. For example, the lake level
will determine the surface area of the lake, which will determine the amount of
incoming shortwave radiation which affects evaporation the overall heat exchane
in the lake. It will also determine the volume of the lake, which will determine
the amount of nutrients in the lake.

Here we demonstrate how to calibrate the lake level for the AEME-GOTM model. It
is important to remember that if there is no observed lake level present in the
`aeme` object, then the lake level will be estimated using supplied rainfall,
inflows (if present), outflows (if present), and calculated evaporation using
air temperature and wind speed and lake surface area. When this cal

```{r build-model, eval = TRUE}
# Load libraries
library(AEME)
library(aemetools)

# Create a temporary directory
tmpdir <- "calib-test"
dir.create(tmpdir, showWarnings = FALSE)
aeme_dir <- system.file("extdata/lake/", package = "AEME")
# Copy files from package into tempdir
file.copy(aeme_dir, tmpdir, recursive = TRUE)
path <- file.path(tmpdir, "lake")
aeme_data <- yaml_to_aeme(path = path, "aeme.yaml")

# Remove observations and outflow from example data
obs <- observations(aeme_data)
obs$lake <- NULL # Remove all observations except lake level
observations(aeme_data) <- obs
outf <- outflows(aeme_data)
outf$data <- NULL # Remove all outflows
outflows(aeme_data) <- outf

mod_ctrls <- read.csv(file.path(path, "model_controls.csv"))
inf_factor = c("dy_cd" = 1, "glm_aed" = 1, "gotm_wet" = 1)
outf_factor = c("dy_cd" = 1, "glm_aed" = 1, "gotm_wet" = 1)
model <- c("gotm_wet")

# Build the ensemble - this includes lake level estimation
aeme_data <- build_ensemble(path = path, aeme_data = aeme_data,
                            model = model, mod_ctrls = mod_ctrls,
                            inf_factor = inf_factor, ext_elev = 5,
                            use_bgc = FALSE, use_lw = TRUE)

```

Let's look at the lake level calculated and the estimated outflows (wbal) for the lake.

```{r plot-level, results = "hold", fig.width = 8, fig.height = 6}
par(mfrow = c(2, 1))
plot(aeme_data, "observations")
plot(aeme_data, "outflows")
```

Now we will run the model and plot the model outputs against the observations.

```{r}
#| label: strategy1-run-plot
#| fig.width: 8
#| fig.height: 6
library(ggplot2)

aeme_data <- run_aeme(aeme_data = aeme_data, model = model,
                      verbose = FALSE, mod_ctrls = mod_ctrls,
                      path = path)
p1 <- plot_output(aeme_data = aeme_data, model = model, var_sim = "HYD_wlev",
                  print_plots = FALSE, ylim = c(12.8, 14)) +
  scale_size_manual(values = c("Obs" = 0.1)) +
  ggtitle("Uncalibrated model")
p1
```

Although, this is a short time period and ideally a longer time period would be
used. A common issue for lakes is to be either gaining too much water or running
out of water. This can be caused by a number of factors:

-   Missing or inaccurate inflow/outflow data
-   Inaccurate measurements of precipitation
-   Missing or inaccurate lake level data
-   Inaccurate estimations of lake volume
-   Inaccurate estimations of lake evaporation

Generally, the most common cause is missing or inaccurate outflow data. 
For this reason, we assume that there is large uncertainty in the outflow data 
and therefore we will calibrate a scaling factor for the outflow data.

```{r include = FALSE}
ggplot2::update_geom_defaults("point", list(size = 1))
```

```{r}
utils::data("aeme_parameters", package = "aemetools")

param <- aeme_parameters[aeme_parameters$model == model, ]
param

```



```{r results = "hide"}
#| label: strategy1-calib
#| fig.width: 8
#| fig.height: 6
#| results: hide
#| message: false

# Function to calculate fitness
fit <- function(O, P) {
  -1 * (cor(x = O, y = P, method = "pearson") -
          (mean(abs(O - P)) / (max(O) - min(O))))
}

ctrl <- list(VTR = -Inf, NP = 20, itermax = 400, reltol = 0.07, p = 0.5,
             mutate = 0.1, parallel = TRUE, out_file = "results.csv",
             na_value = 999)

vars_sim <- c("HYD_wlev")
weights <- c("HYD_wlev" = 10)

# Calibrate AEME model
ctrl <- calib_aeme(aeme_data = aeme_data, path = path,
                   param = param, model = model,
                   mod_ctrls = mod_ctrls, FUN = fit, ctrl = ctrl,
                   vars_sim = vars_sim, weights = weights)

calib_res <- read_calib(ctrl = ctrl, model = model)

testthat::expect_true(is.data.frame(calib_res))

plist <- plot_calib(calib = calib_res, model = model,
                    na_value = ctrl$na_value)
plist$dotty

# library(patchwork)
# wrap_plots(plist, guides = "collect", design = "AB
#            CC")

```

Extract the calibrated parameters and run the model again.

```{r}
#| label: strategy1-run-calib-plot
#| fig.width: 8
#| fig.height: 6
#| results: hide

# Extract the calibrated parameters
best_pars <- get_param(calib = calib_res, model = model, 
                       na_value = ctrl$na_value, best = TRUE)

param <- update_param(param = param, best_pars = best_pars)

aeme_data <- run_aeme_param(aeme_data = aeme_data, model = model, 
                            mod_ctrls = mod_ctrls, path = path, param = param,
                            return_aeme = TRUE)

p2 <- plot_output(aeme_data = aeme_data, model = model, var_sim = "HYD_wlev",
                  print_plots = FALSE, ylim = c(12.8, 14)) +
  scale_size_manual(values = c("Obs" = 0.1)) +
  ggtitle("Calibrated model")

library(patchwork)
p1 + p2
```

