---
title: "Calibrate AEME"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrate AEME}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
library(printr)
unlink("calib-test", recursive = TRUE, force = TRUE)
```

## Setup

First, we will load the `AEME` and `aemetools` package:

```{r setup, message = FALSE, warning=FALSE}
library(AEME)
library(aemetools)
```

Create a folder for running the example calibration setup.

```{r create-folder}

# tmpdir <- "calib-test"
# dir.create(tmpdir, showWarnings = FALSE)
tmpdir <- tempdir()
aeme_dir <- system.file("extdata/lake/", package = "AEME")
# Copy files from package into tempdir
file.copy(aeme_dir, tmpdir, recursive = TRUE)
path <- file.path(tmpdir, "lake")

list.files(path, recursive = TRUE)
```

## Build AEME ensemble

Using the `AEME` functions, we will build the AEME model setup. For this
example, we will use the `glm_aed` model. The `build_aeme` function will build the model setup and the `run_aeme` function will run the model. The model output is stored in the lake folder.

Note that `AEME` functions can be used in pipelines to allow for more efficient code. For example, the `build_aeme` function can be used in a pipeline with the `run_aeme` function to build and run the model in one step.

```{r build-aeme-example, warning=FALSE, message=FALSE}

aeme <- yaml_to_aeme(path = path, "aeme.yaml")
model_controls <- AEME::get_model_controls()
model <- c("glm_aed")
aeme <- build_aeme(path = path, aeme = aeme, model = model, 
                   model_controls = model_controls, ext_elev = 5,
                   use_bgc = FALSE) |> 
  run_aeme()

```



```{r}
#| label: plot-aeme-example
#| fig.height: 4
#| fig.width: 8
#| cache: false
#| fig.cap: Water temperature contour plotfor the model output.

plot(aeme)
```

## Load parameters to be used for the calibration

Parameters are loaded from the `aemetools` package within the `aeme_parameters`
dataframe. The parameters are stored in a data frame with the following columns:

-   `model`: The model name

-   `file`: The file name of the model parameter file. For meteorological
    scaling variables, "met" is used, whereas for scaling factors for inflow and
    outflow, "inflow" an "outflow" is used accordingly.

-   `name`: The parameter name

-   `value`: The parameter value

-   `min`: The minimum value of the parameter

-   `max`: The maximum value of the parameter

-   `module`: The module of the parameter

-   `group`: The group of the parameter. Only used for phytoplankton and
    zooplankton parameters.

Parameters to be used for the calibration.

```{r}
#| label: load-parameters
#| tbl-captions: "Parameters to be used for the calibration."
utils::data("aeme_parameters", package = "AEME")
aeme_parameters|>
  DT::datatable(options = list(pageLength = 4, scrollX = TRUE))

```

This dataframe can be modified to change the parameter values. For example, we
can change the `light/Kw` parameter for the `glm_aed` model to 0.1:

```{r change-parameters}
aeme_parameters[aeme_parameters$model == "glm_aed" &
                  aeme_parameters$name == "light/Kw", "value"] <- 0.1
aeme_parameters
```

This dataframe can be passed to the `run_aeme_param` function to run AEME with
the parameter values specified in the dataframe. This function is different to
the `run_aeme` function in that it does not return an `aeme` object, but the
model output is generated within the lake folder.

```{r run-aeme-param}
run_aeme_param(aeme = aeme, param = aeme_parameters,
                 model = model, path = path)

```

## Calibration setup

### Choosing variables to calibrate 

Choosing which variables to calibrate is an important step in the calibration
process. The variables to calibrate are usually selected based on the 
availability of data and the importance of the variable to the model. 

There is a function within the `AEME` package called `get_mod_obs_vars` which 
can be used to get the available variables for which there is modelled and observed data.

```{r select-calib-vars}
available_vars <- AEME::get_mod_obs_vars(aeme = aeme, model = model)
available_vars
```

There are 10 variables available for calibration, this includes derived 
variables such as thermocline depth (HYD_thmcln) and Schmidt stability 
(HYD_schstb).

For this example, we will calibrate the water temperature and lake level. The
variables are selected using the AEME variable definition e.g.
`c("HYD_temp", "LKE_lvlwtr")`.

```{r define-variables}
vars_sim <- c("HYD_temp", "LKE_lvlwtr")
```


### Define fitness function

First, we will define a function for the calibration function to use to
calculate the fitness of the model. This function takes a dataframe as an
argument. The dataframe contains the observed data (`obs`) and the modelled data
(`model`). The function should return a single value.

Here we use the root mean square error (RMSE) as the fitness function:

$$\text{RMSE}(y, \hat{y}) = \sqrt{\frac{\sum_{i=0}^{N - 1} (y_i - \hat{y}_i)^2}{N}}$$

```{r define-fitness-function}
# Function to calculate fitness
rmse <- function(df) {
  sqrt(mean((df$obs - df$model)^2))
}

```

Different functions can be applied to different variables. For example, we can
use the RMSE for the lake level and the mean absolute error (MAE) for the water
temperature:

```{r define-fitness-function-2}
# Function to calculate fitness
mae <- function(df) {
  mean(abs(df$obs - df$model))
}

```

Then these would be combined into a named list of functions which will be passed
to the `calib_aeme` function. They are named according to the target variable.

```{r define-function-list}
# Create list of functions
FUN_list <- list(HYD_temp = mae, LKE_lvlwtr = rmse)

```

### Define control parameters

Next, we will define the control parameters for the calibration. The control
parameters are generated using `create_calib_control()` and are then passed to
`calib_aeme()`. (`create_control()` still exists but is deprecated in favour of
the dedicated `create_calib_control()` and `create_sa_control()` constructors.)
The control parameters for calibration are as follows:

```{r create-control-args, tidy=FALSE, printr.help.sections=c('arguments'), comment=''}
?create_calib_control
```

Here is an example of the control parameters for calibration, with a
value-to-reach of 0, 40 members in each population, maximum number of iterations
of 400, a relative tolerance of 0.07, 25% of parameters in each population need
to be non-NA to be used as parents in the next generation, 10% of the children
parameters undergo random mutation, parallel processing is used, the file type
for writing the results is CSV, the NA value is 999 and 2 cores are used for
parallel processing. The control parameters are stored in the `ctrl` object
which is then passed to the `calib_aeme` function.

```{r define-control-parameters}
ctrl <- create_calib_control(VTR = 0, NP = 40, itermax = 400, reltol = 0.07, 
                             cutoff = 0.25, mutate = 0.1, parallel = TRUE,
                             file_type = "csv", na_value = 999, ncore = 2)
```

### Define variable weights

Weights need to be attributed to each of the selected variables. The weights are
used to scale the fitness value. This can be helpful if the variables have 
different units. For example, if the temperature is in
degrees Celsius and the water level is in metres, then the water level will have
a much larger impact on the fitness value. Therefore, the weight for the water
level should be much smaller than the weight for the temperature.

The weights are specified in a named vector. The names of the vector should be
the same as the variable names.

```{r define-variable-weights}
weights <- c("HYD_temp" = 1, "LKE_lvlwtr" = 0.1)
```

## Run calibration

Once we have defined the fitness function, control parameters and variables, we
can run the calibration. The `calib_aeme` function takes the following
arguments:

```{r calib-aeme-args, tidy=FALSE, printr.help.sections=c('arguments'), comment=''}
?calib_aeme
```

The `calib_aeme` function writes the calibration results to the file specified
after each generation. This allows the calibration to be stopped and restarted
at any time. The `calib_aeme` function returns the `ctrl` object with any
updated values.

```{r run-calibration, cache=FALSE, results='hide'}
sim_id <- calib_aeme(aeme = aeme, path = path,
                     param = aeme_parameters, model = model,
                     FUN_list = FUN_list, ctrl = ctrl, 
                     vars_sim = vars_sim, weights = weights)

```

## Visualise calibration results

The calibrations results can be read in using the `read_calib` function. This
function takes the following arguments:

```{r read-calib-args, tidy=FALSE, printr.help.sections=c('arguments'), comment=''}
?read_calib
```

The `read_calib` function returns a dataframe with the calibration results. The
calibration results include the model, generation, index (model run), parameter
name, parameter value, fitness value and the median fitness value for each
generation.

These results can be visualised using the `plot_calib` function. This function
takes the following arguments:

-   `calib`: The calibration results as read in using the `read_calib` function.
-   `model`: The model used for the calibration.
-   `ctrl`: The control parameters used for the calibration.

And returns a list of ggplot objects: a dotty plot, density plot and convergence
plot.

```{r visualise-calibration-results, cache = FALSE}

calib <- read_calib(ctrl = ctrl, sim_id = sim_id)
plist <- plot_calib(calib = calib)
```

### Dotty plot

This can be used for comparing sensitivity across parameters. The dotty plot
shows the fitness value for each parameter value for each generation. The
fitness value is on the y-axis and the parameter value is on the x-axis. It is
faceted by the parameter name. The parameter values are coloured by the
generation. The best fitness value for each generation is shown as a black line
with a red dot.

```{r dotty-plot, fig.width = 8, fig.height = 6}
plist$dotty
```

### Histogram plot

This is useful for comparing the distribution of parameter values across
generations. The histogram plot shows the frequency of the parameter values for
each generation. The parameter values are on the x-axis and the density is on
the y-axis. It is faceted by the parameter name.

If a parameter is converging on a value, then the histogram will show a peak
around that value. If a parameter is not converging on a value, then the
histogram will show a flat distribution.

```{r histogram-plot, fig.width = 8, fig.height = 6}
plist$histogram
```

### Convergence plot

This is more generally used for assessing model convergence. The convergence
plot shows the values use over the iterations. The parameter value is on the
y-axis and the iteration is on the x-axis. It is faceted by the parameter name.
The parameter values are coloured by the generation. The best fitness value for
each generation is shown as a solid horizontal black line.

```{r convergence-plot, fig.width = 8, fig.height = 6}
plist$convergence
```

## Assess calibrated values

The single best parameter set is extracted with `get_best_params()`, which takes:

-   `calib`: the calibration results as read in with `read_calib()`.
-   `fit_col`: the name of the column holding the fitness value. Defaults to
    `"fit"`.
-   `na_value`: the sentinel used for crashed runs, dropped before ranking.
    Defaults to the value stored on the control.

Use `get_sim_params(calib, quantile_threshold = )` for every evaluated set below
a fitness quantile rather than just the best one, and `update_param()` to fold a
calibrated set back into a `param` dataframe. (`get_param()` is retained as a
thin wrapper over these two.)

```{r extract-best-parameters}
best_params <- get_best_params(calib)
best_params

```

The best parameter values can be used to run the model and compare the simulated
values to the observed values. This can be done using the `run_aeme_param`
function.

```{r run-calibrated-model}
aeme <- run_aeme_param(aeme = aeme, path = path,
                       param = best_params, model = model,
                       return_aeme = TRUE)

```

The simulated values can be compared to the observed values using the
`assess_model` function. This function takes the following arguments:

-   `aeme`: The `aeme` object which has observations and model simulations.
-   `model`: The model to assess.
-   `var_sim`: The variables to use for the assessment.

The `assess_model` function returns:

```{r sa-aeme-args, tidy=FALSE, printr.help.sections=c('value'), comment=''}
?assess_model
```

```{r assess-calibated-model}

assess_model(aeme = aeme, model = model, var_sim = vars_sim)

```

### Reproducing the calibration fit value

`assess_model()`'s table above is a general-purpose diagnostic: a fixed set of
per-variable, unweighted statistics (`nse`, `kge`, `mae`, ...) computed however
`var_sim` happens to be assessed right now. It is **not** the number
`calib_aeme()` recorded as `best_params$fit_value` for this parameter set, and
is not expected to match it - even run on the exact best parameters. That
number is the calibration objective itself: whatever `FUN_list` returned,
weighted per variable and summed across `vars_sim` (this is what
`eval_param_chunk()` does internally on every candidate). To reproduce it for
verification, call `run_and_fit()` again with the *same* `param`, `FUN_list`,
`vars_sim` and `weights` used for the original `calib_aeme()` call, then
reduce it the same way `eval_param_chunk()` does:

```{r verify-calib-fit}
res <- run_and_fit(aeme = aeme, param = best_params, model = model,
                   vars_sim = vars_sim, path = path, FUN_list = FUN_list,
                   weights = weights,
                   include_wlev = "LKE_lvlwtr" %in% vars_sim,
                   method = "calib", fit = TRUE)
fit <- if (any(is.na(unlist(res)))) ctrl$na_value else sum(unlist(res))
fit
min(best_params$fit_value)
```

`fit` matches `best_params$fit_value`; `res` holds the weighted value for each
entry of `vars_sim` before summation. Use this pattern to check a recorded fit
value is reproducible, and `assess_model()` separately for interpretable,
per-variable model performance - they answer different questions and are not
interchangeable.

(One caveat if a model runs sub-daily: `run_and_fit()` and
`assess_aeme()`/`assess_model()` extract modelled values through different
`AEME` code paths and can pick a different output timestep for the same
observation day, which can shift the comparison slightly even beyond the
differences described above. Daily-output runs, as here, are unaffected.)

### Visualise model performance

The model performance can be visualised using the `plot_resid` function within
the AEME package. This returns a list of ggplot objects, a plot of residuals for
each variable. This is a multi-panel plot displaying residuals for:

-   Observed vs. predicted values
-   Residuals vs. predicted values
-   Residuals vs. day of year
-   Residuals vs. quantiles of the observed values

```{r plot-residuals}
pl <- lapply(vars_sim, \(v) {
  plot_resid(aeme = aeme, model = model, var_sim = v)
})
names(pl) <- vars_sim

```

#### Water temperature residuals

```{r plot-residuals-plot, fig.width = 8, fig.height = 6}

pl$HYD_temp
  
```

#### Lake level residuals

```{r plot-residuals-plot-2, fig.width = 8, fig.height = 6}

pl$LKE_lvlwtr

```

## Other calibration approaches

The workflow above uses the built-in differential-evolution engine. `aemetools`
also provides:

-   **Multi-objective (Pareto) calibration** with the `"MOEDA"` method. Set
    `create_calib_control(c_method = "MOEDA")` and pass a parameter covariance
    matrix built with `create_param_var_matrix()` to
    `calib_aeme(param_var_matrix = )` (the two are mutually required). Extract
    the non-dominated sets with `get_pareto_front()` and inspect them across
    generations with `plot_pareto_generations()`.
-   **PEST++ calibration** using the `pestpp-ies` iterative ensemble smoother,
    with Morris screening and localisation. Install the solver once with
    `install_pest()`, then see `vignette("screen-then-calibrate")` and
    `vignette("staged-calibration")`.
-   **Split-sample validation**: `get_calib_periods()` proposes calibration and
    validation windows from the observation record (or returns the whole record
    with `split = FALSE`), and `validate_aeme()` scores a parameter set over the
    held-out period.
-   A `timeout` argument on `create_calib_control()` returns the best set found
    so far if a wall-clock limit is reached.
