---
title: "Download ERA5 data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Download ERA5 data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

ecmwfr::wf_set_key(key = Sys.getenv("CDS_KEY"),
                   user = Sys.getenv("CDS_USER"))

```

```{r include=FALSE, echo=FALSE}
library(aemetools)

data("era5_ref_table", package = "aemetools")
era5_names <- paste0(tolower(era5_ref_table$variable[-1]), collapse = ", ")

```

```{r load-libraries, warning=FALSE, message=FALSE, results='hide'}
library(aemetools)
library(sf)
library(terra)
library(tmap)
tmap_mode("view")
tmap_options(basemap.server = "OpenTopoMap")

```


## Global

### Download global point meteorological data

One of the most widely used meteorological datasets for hydrodynamic modelling
is the [ERA5-Land reanalysis](https://cds.climate.copernicus.eu/datasets/reanalysis-era5-land?tab=overview)
dataset. This dataset has a spatial resolution of \~9 km and provides hourly
meteorological data for the period 1950-present. It also includes a wide range
of meteorological variables that are required to drive hydrological and
hydrodynamic models.

However, there are known issues with the ERA5-Land dataset, such as the
underestimation of precipitation and wind speed in some regions. Therefore, it
is recommended to compare the ERA5-Land data with local observations before
using it for hydrodynamic modelling.

#### Set up Copernicus (CDS) account and link with `ecmwfr` package

Please see instructions on setting up a Copernicus (CDS) account and linking it
with the `ecmwfr` package in the [ecmwfr vignette](https://bluegreen-labs.github.io/ecmwfr/#use).

```{r setup-cds-key, eval=FALSE}
key <- "XXXXXXXX-XXXX-XXXX-XXXX-XXXXXXXXXXXX"
user <- "person@email.com"

# Add the key and user to the environment
Sys.setenv(CDS_KEY = key)
Sys.setenv(CDS_USER = user)

# Set the key for connecting to Copernicus Data Store (CDS)
ecmwfr::wf_set_key(key = Sys.getenv("CDS_KEY"),
                   user = Sys.getenv("CDS_USER"))

```

### Download ERA5 data

#### Point timeseries data

The `get_era5_isimip_point()` function can be used to download ERA5 
meteorological data for a specific point location. This uses the ISIMIP3a
dataset which is a subset of the ERA5 dataset and is accessed via the 
[ISIMIP API](https://files.isimip.org/api/v2). This function requires the 
latitude, longitude, years, and variables to be specified. It can be used for 
any point location globally. However, this dataset only covers 1900-2021.

##### Example: Download meterological data for 2021 for Lake Toba, Indonesia

However, it is very quick and easy to download meteorological data for a 
specific point location. We will choose the latitude and longitude of Lake Toba,
Indonesia, and download the air temperature and precipitation for the year 2021.

```{r}
lon <- 98.67591
lat <- 2.637047
years <- 2021
vars <- c("MET_tmpair", "MET_pprain")

met <- get_era5_isimip_point(lon, lat, years, vars)
summary(met)

```



#### GRIB files

The `download_era5_grib()` function can be used to download ERA5 meteorological
data from the Copernicus Data Store (CDS). This function requires the latitude,
longitude, variable, year, and month to be specified. The data will be saved as
a GRIB file in the specified path.

##### Example: Download 2m air temperature GRIB files for January 2024 for Lake Toba, Indonesia

We will choose the latitude and longitude of Toba, Indonesia, and download the
2m temperature for January 2024.

#### Download shapefile for Lake Toba, Indonesia

We will download a shapefile for Lake Toba, Indonesia, and plot the location
using the `osmdata` and `tmap` packages.

For more details on using `osmdata`, you can visit the [osmdata website](https://docs.ropensci.org/osmdata/).

```{r download-toba}
library(osmdata)
osm_data <- opq(bbox = "Sumatra, Indonesia") |> 
  add_osm_feature(key = "name:en", value = "Lake Toba", value_exact = FALSE) |> 
  osmdata_sf()

# Extract lake polygon
toba <- osm_data$osm_multipolygons |> 
  st_make_valid()
toba

```
Plot the location of Lake Toba.

```{r plot-toba}
tm_shape(toba) +
  tm_fill(fill = "blue", fill_alpha = 0.5) +
  tm_borders(col = "black") 
```

Alternatively, you can manually input the latitude and longitude of the location
you are interested in.

```{r plot-location}
lat <- 2.637047
lon <- 98.67591

# View location
coords <- data.frame(lat = lat, lon = lon) |> 
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

tm_shape(coords) +
  tm_dots(fill = "red", size = 2) 

```

Now we can download the ERA5 data for the 2m temperature for January 2024. We 
can use the `download_era5_grib()` function to download the data using the 
shapefile we downloaded earlier or the latitude and longitude of the location.


```{r results='hide', message=FALSE, eval=FALSE}
year <- 2024
month <- 1
variable <- "2m_temperature"
path <- "data/test"
site <- "toba"

files <- download_era5_grib(shape = toba, year = year, month = month,
                            variable = variable, path = path, site = site,
                            user = user)
```

```{r results='hide', message=FALSE, eval=TRUE, include=FALSE}
path <- "data/test"
site <- "toba"
era5_dir <- system.file("extdata/era5/", package = "aemetools")
dir.create(path = path, recursive = TRUE, showWarnings = FALSE)
fils <- list.files(era5_dir, full.names = TRUE, pattern = site)
file.copy(fils, path, recursive = TRUE)
files <- list.files(path, full.names = TRUE, pattern = site)
```


The function will return the path to the downloaded file.

```{r}
files
```
Files are in GRIB (GRIdded Binary) format, which is a format commonly used in
meteorology to store gridded data. You can read more about these files on the
ECMWF website: [GRIB](https://confluence.ecmwf.int/display/CKB/What+are+GRIB+files+and+how+can+I+read+them).

The most important thing is not to be afraid of them. You can easily read them
using the `terra::rast()` function.

```{r}
r <- rast(files[1])
r
```

Hmmmmmmm... Ok, maybe these files are not that easy to read/understand. We can 
also plot them in leaflet using the `terra::plet()` function.

```{r}
plet(r)
```

As you can see it gives us a spatial grid of the 2m temperature for the month of
January 2024. However it is at a 1-hour temporal resolution. We can use the
`read_grib_point()` function to extract the timeseries data at a specific point 
location. 

```{r}
df <- read_grib_point(file = files, shape = toba)
head(df)
```



## New Zealand

### Download NZ point meteorological data

Currently, there are 
[ERA5-Land data](https://cds.climate.copernicus.eu/datasets/reanalysis-era5-land?tab=overview) 
(\~9km grid spacing) archived for New Zealand (166.5/-46.6/178.6/-34.5)
for the time period 1980-2023 with the main meteorological variables
(`r era5_names`) required to drive hydrological and hydrodynamic models. This
can be easily downloaded using the example below. 

```{r setup}
library(aemetools)
```
First we will define our point location for New Zealand. We will use the
latitude and longitude of Lake Rotorua, Bay of Plenty, New Zealand.

```{r define-location}
lon <- 176.2717
lat <- -38.079
```

Plot the location.

```{r}
coords <- data.frame(lat = lat, lon = lon) |> 
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

tm_shape(coords) +
  tm_dots(fill = "red", size = 2) 


```

Now we can download the ERA5 data for the air temperature and precipitation for
the years 2000-2001. We can use the `get_era5_land_point_nz()` function to download the
data.

The function will return a dataframe with the daily meteorological data for the
specified years.

```{r download-era5}

vars <- c("MET_tmpair", "MET_pprain")

met <- get_era5_land_point_nz(lat = lat, lon = lon, years = 2000:2001,
                              vars = vars)
summary(met)
```

We can plot the timeseries data using the `ggplot2` package.

```{r plot-era5, fig.alt='Timeseries plot of air temperature and precipitation.'}

library(ggplot2)
library(tidyr)

met |> 
  pivot_longer(cols = c(MET_tmpair, MET_pprain)) |> 
  ggplot(aes(x = Date, y = value)) +
  geom_line() +
  facet_wrap(~name, scales = "free_y", ncol = 1) +
  theme_bw()

```

By default, it will download all available variables. Currently available
variables are air temperature (`MET_tmpair`), dew point temperature
(`MET_tmpdew`), wind u-component (`MET_wnduvu`), wind v-component (`MET_wnduvv`),
precipitation (`MET_pprain`), snowfall (`MET_ppsnow`), surface level pressure (`MET_prsttn`)
and shortwave radiation (`MET_radswd`).

```{r tbl-met-vars, results='asis', include=FALSE}

era5_ref_table |> 
  dplyr::select(variable, aeme, nc_unit) |>
  dplyr::rename(Variable = variable, Aeme = aeme, Unit = nc_unit) |>
  dplyr::mutate(Unit = dplyr::case_when(
    Aeme %in% c("MET_radswd", "MET_radlwd") ~ "W/m²",
    .default = Unit
  )) |> 
  knitr::kable()

```

To download all available variables for Lake Rotorua for the years 2022-2023,
you can use the following code:
```{r download-era5-all}
met <- get_era5_land_point_nz(lat = lat, lon = lon, years = 2022:2023)
summary(met)
```

## Derive other meteorological variables

The downloaded ERA5 data can be converted to other meteorological variables
using the `AEME::expand_met()` function. Which will calculate the following
variables:

- `MET_tmpdew`: Dew point temperature (°C)
- `MET_wndspd`: Wind speed (m/s) 
- `MET_wnddir`: Wind direction (°)
- `MET_radlwd`: Longwave radiation (W/m²)
- `MET_cldcvr`: Cloud cover (fraction)
- `MET_prmslp`: Mean sea level pressure (Pa)
- `MET_humrel`: Relative humidity (%)
- `MET_prvapr`: Vapour pressure (Pa)

To use this function, you will need to provide the latitude, longitude, and
elevation of the site. Elevation is required to calculate the mean sea level
pressure and cloud cover. 

We can access the elevation data using the `get_dem_value()` function. This 
function uses the LINZ data service to obtain the elevation value for a
given latitude and longitude using the national 8m digital elevation model (DEM)
layer (ID: 51768).

This requires a LINZ API key which can be obtained by visiting the
[LINZ website](https://data.linz.govt.nz/layer/51768-nz-8m-digital-elevation-model-2012/) 
and registering for an account. Then adding it to your R session using the
`add_linz_key()` function.

```{r get-dem}

elev <- get_dem_value(lat = lat, lon = lon)
elev

```

Now we can expand the meteorological data to include the derived variables.

```{r expand-era5}
met <- AEME::expand_met(met, lat = lat, lon = lon, elev = elev)
summary(met)

```

With these simple steps, you can download and derive meteorological data for
your location of interest. This data can be used to drive hydrological and
hydrodynamic models for your study area.

Head over to the [AEME Getting Started](https://limnotrack.github.io/AEME/articles/getting-started.html) vignette to learn more about how to use the meteorological data to drive hydrodynamic models.

