---
title: "Download ERA5 data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Download ERA5 data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r include=FALSE, echo=FALSE}
library(aemetools)

data("era5_ref_table", package = "aemetools")
era5_names <- paste0(tolower(era5_ref_table$variable[-1]), collapse = ", ")

```

## Global

### Download global point meteorological data

One of the most widely used meteorological datasets for hydrodynamic modelling
is the [ERA5-Land reanalysis](https://cds.climate.copernicus.eu/datasets/reanalysis-era5-land?tab=overview)
dataset. This dataset has a spatial resolution of \~9 km and provides hourly
meteorological data for the period 1950-present. It also includes a wide range
of meteorological variables that are required to drive hydrological and
hydrodynamic models.

However, there are known issues with the ERA5-Land dataset, such as the
underestimation of precipitation and wind speed in some regions. Therefore, it
is recommended to compare the ERA5-Land data with local observations before
using it for hydrodynamic modelling.

#### Set up Copernicus (CDS) account and link with `ecmwfr` package

Please see instructions on setting up a Copernicus (CDS) account and linking it
with the `ecmwfr` package in the [ecmwfr vignette](https://bluegreen-labs.github.io/ecmwfr/#use).

```{r setup-cds-key, eval=FALSE}
Sys.setenv(CDS_KEY = "XXXXXXXX-XXXX-XXXX-XXXX-XXXXXXXXXXXX")
Sys.setenv(CDS_USER = "person@email.com")
```

```{r setup-cds-key-2, echo=FALSE, results='hide'}
ecmwfr::wf_set_key(key = Sys.getenv("CDS_KEY"),
                   user = Sys.getenv("CDS_USER"))
```

```{r results='hide'}
lat <- 2.637047
lon <- 98.67591
year <- 2024
month <- 1
variable <- "2m_temperature"
path <- "data/test"
files <- download_era5_grib(lat = lat, lon = lon, year = year, month = month,
                            variable = variable, path = path, site = "toba",
                            user = Sys.getenv("CDS_USER"))
df <- read_grib_point(file = files, lat = lat, lon = lon)
```

```{r}
head(df)
```



## New Zealand

### Download NZ point meteorological data

Currently, there are 
[ERA5-Land data](https://cds.climate.copernicus.eu/datasets/reanalysis-era5-land?tab=overview) 
(\~9km grid spacing) archived for New Zealand (166.5/-46.6/178.6/-34.5)
for the time period 1980-2023 with the main meteorological variables
(`r era5_names`) required to drive hydrological and hydrodynamic models. This
can be easily downloaded using the example below. 

```{r setup}
library(aemetools)
```

```{r download-era5}
lon <- 176.2717
lat <- -38.079
variables <- c("MET_tmpair", "MET_pprain")

met <- get_era5_point(lat = lat, lon = lon, years = 2000:2001,
                      variables = variables)
summary(met)
```

```{r plot-era5}

library(ggplot2)
library(tidyr)

met |> 
  pivot_longer(cols = c(MET_tmpair, MET_pprain)) |> 
  ggplot(aes(x = Date, y = value)) +
  geom_line() +
  facet_wrap(~name, scales = "free_y", ncol = 1) +
  theme_bw()

```

By default, it will download all available variables.

```{r download-era5-all}
lon <- 176.2717
lat <- -38.079

met <- get_era5_point(lat = lat, lon = lon, years = 2022:2023)
summary(met)
```

## Derive other meteorological variables

The downloaded ERA5 data can be converted to other meteorological variables
using the `AEME::expand_met()` function. Which will calculate the following
variables:

- `MET_tmpdew`: Dew point temperature (°C)
- `MET_wndspd`: Wind speed (m/s) 
- `MET_wnddir`: Wind direction (°)
- `MET_radlwd`: Longwave radiation (W/m²)
- `MET_cldcvr`: Cloud cover (fraction)
- `MET_prmslp`: Mean sea level pressure (Pa)
- `MET_humrel`: Relative humidity (%)
- `MET_prvapr`: Vapour pressure (Pa)

To use this function, you will need to provide the latitude, longitude, and
elevation of the site. Elevation is required to calculate the mean sea level
pressure and cloud cover. 

We can access the elevation data using the `get_dem_value()` function. This 
function uses the LINZ data service to obtain the elevation value for a
given latitude and longitude using the national 8m digital elevation model (DEM)
layer (ID: 51768).

This requires a LINZ API key which can be obtained by visiting the
[LINZ website](https://data.linz.govt.nz/layer/51768-nz-8m-digital-elevation-model-2012/) 
and registering for an account. Then adding it to your R session using the
`add_linz_key()` function.

```{r get-dem}

elev <- get_dem_value(lat = lat, lon = lon)
elev

```


 




```{r expand-era5}
met <- AEME::expand_met(met, lat = lat, lon = lon, elev = elev)
summary(met)

```
