---
title: "Using the Limnotrack API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the Limnotrack API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: setup
#| echo: FALSE
#| eval: TRUE

library(AEME)
library(aemetools)
library(tmap)
tmap_mode("view")
```

The Limnotrack API provides access to various environmental data layers, including
meteorological data from ERA5-Land. This vignette demonstrates how to use the
Limnotrack API to extract ERA5-Land data for specific locations in New Zealand.

First, it is always good to check the status of the API to ensure it is 
operational.

```{r}
#| label: check-api
#| echo: TRUE
#| eval: TRUE

check_api_status()

```


## Lake shape data

There are shape files for all lakes in the LimnoTrack database. These can be
accessed using the `get_lake_shape()` function. You need to provide the lake ID
(e.g. FENZ ID or Lernzmp ID).

```{r}
#| label: get-lake-shape
#| echo: TRUE
#| eval: TRUE

lake <- get_lake_shape(id = 15022)
print(lake)
```
```{r}
#| label: plot-lake-shape
#| echo: TRUE
#| eval: TRUE

tm_shape(lake) +
  tm_borders(col = "blue", lwd = 2) +
  tm_basemap("Esri.WorldImagery")
```


## Lake catchment data

Lake catchment data can be accessed using the `get_catchment_data()` function. You
need to provide the lake ID (e.g. FENZ ID or Lernzmp ID). The downloaded data
is a list which includes the catchment, reaches, lakes, subcatchments, and LCDB
data.

```{r}
#| label: get-catchment-data
#| echo: TRUE
#| eval: TRUE

catchment <- get_catchment_data(id = 15022)
names(catchment)
```

```{r}
#| label: plot-catchment-data
#| echo: TRUE
#| eval: TRUE

tm_shape(catchment$catchment) +
  tm_borders(col = "red", lwd = 2) +
  tm_shape(catchment$reaches) +
  tm_lines(col = "reachtype2") +
  tm_shape(catchment$lakes) +
  tm_borders(col = "cyan", lwd = 2) +
  tm_shape(catchment$subcatchments) +
  tm_borders(col = "green", lwd = 2) +
  tm_shape(catchment$lcdb) +
  tm_fill(id = "Name_2018", fill = "Name_2018", fill_alpha = 0.5) 
  # tm_basemap("Esri.WorldImagery")
```

## Aeme object

For many lakes in the LimnoTrack database, there is an associated Aeme object
that contains model parameters and other information. You can access the Aeme
object using the `get_aeme()` function. You need to provide the lake ID
(e.g. FENZ ID or Lernzmp ID).



```{r}
#| label: get-aeme
#| echo: TRUE
#| eval: TRUE

aeme <- get_aeme(id = 15022)
aeme
```

You can plot the lake hypsograph from the Aeme object using the `plot_hyps()`
function.

```{r}
#| label: plot-aeme-hyps
#| echo: TRUE
#| eval: TRUE

plot_hyps(aeme, y = "depth")
```
