---
title: "Run Hydrological models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run Hydrological models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(aemetools)
```

### Hydrological modelling - Run GR4J model

Here is simple example set up for one of the lake inflows into Lake Rotorua. First, the input for the model are generated using the stream ID (nzsegment), and spatial features (sf objects) of the reaches, lake and catchment (including sub-catchments), observed discharge (if available) meteorological data (air temperature and precipitation). It recursively creates an upstream network using the nzsegment, then combines the subcatchments of all the upstream reaches (`sf::st_union()`) to calculate the area of the catchment.

```{r make-gr4j-inputs, results='asis', fig.keep='all', warning=FALSE}
lat <- -38.079
data_dir <- system.file("extdata/hydro/", package = "aemetools")
lake <- readRDS(file.path(data_dir, "lake.rds"))
reaches <- readRDS(file.path(data_dir, "reaches.rds"))
catchments <- readRDS(file.path(data_dir, "catchments.rds"))
met <- readRDS(file.path(data_dir, "met.rds"))
obs_flow <- readRDS(file.path(data_dir, "obs_flow.rds"))
FUN_MOD <- airGR::RunModel_GR4J
id <- 4087861 # nzsegment

inputs <- make_GR_inputs(id = id, reaches = reaches, lake = lake,
                         catchments = catchments, obs_flow = obs_flow, met = met,
                         lat = lat, FUN_MOD = FUN_MOD,
                         plot = TRUE)


```

Within the `airGR` package, there are calibration algorithms which allows you to calibrate the hydrological model if discharge data for the reach is available. The calibrated parameters can be passed to the `run_GR` function to run the selected model.

```{r run-gr4j, warning=FALSE}
#' airGR uses indices to run the model, so first we split our observed data in 
#' half (0.5) for calibration and validation periods based on when the 
#' observation data starts (which is provided in `inputs$data$start`).
idx_spl <- floor(nrow(inputs$data[inputs$start:nrow(inputs$data), ])
                 * 0.5)
#' Use a model warmup period as everything before when the observations start.
warmup <- 1:(inputs$start - 1)

# Set the indices for the calibration period
cal_idx <- inputs$start:(idx_spl + inputs$start)

# Run the calibration and assign the output
calib <- calib_GR(inputs = inputs, warmup = warmup, run_index = cal_idx)

# Extract the calibrated parameters
param <- calib$ParamFinalR

# Run the model
output <- run_GR(inputs = inputs, param = param,
                 warmup = warmup, run_index = cal_idx)

# Plot the output
plot(output, Qobs = inputs$data$Qmm[cal_idx])
```
