---
title: "Run Hydrological models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run Hydrological models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(aemetools)
```

### Hydrological modelling - Run GR4J model

Here is simple example set up for one of the lake inflows into Lake Rotorua. First, the input for the model are generated using the stream ID (nzsegment), and spatial features (sf objects) of the reaches, lake and catchment (including sub-catchments), observed discharge (if available) meteorological data (air temperature and precipitation). It recursively creates an upstream network using the nzsegment, then combines the subcatchments of all the upstream reaches (`sf::st_union()`) to calculate the area of the catchment.

```{r make-gr4j-inputs, results='asis', fig.keep='all', warning=FALSE}
lat <- -38.079
data_dir <- system.file("extdata/hydro/", package = "aemetools")
lake <- readRDS(file.path(data_dir, "lake.rds"))
reaches <- readRDS(file.path(data_dir, "reaches.rds"))
catchments <- readRDS(file.path(data_dir, "catchments.rds"))
met <- readRDS(file.path(data_dir, "met.rds"))
obs_flow <- readRDS(file.path(data_dir, "obs_flow.rds"))
id <- 4087861 # nzsegment

hydro_model <- make_GR_inputs(id = id, reaches = reaches, lake = lake,
                               catchments = catchments, obs_flow = obs_flow,
                               met = met, lat = lat,
                               FUN_MOD = airGR::RunModel_GR4J,
                               plot = TRUE)
hydro_model
```

`make_GR_inputs()` returns a `HydroModel` S4 object. Its key slots are:

| Slot | Description |
|------|-------------|
| `@inputs_model` | `airGR` `InputsModel` object |
| `@data` | Combined meteorological and observed flow data |
| `@start` | Row index where the observation period begins |
| `@fun_mod` | `airGR` model function |
| `@fun_mod_name` | Display name of the model function |
| `@catchment_area` | Total catchment area in m² |
| `@dem` | Optional DEM raster (default `NULL`) |
| `@land_cover` | Optional land cover data (default `NULL`) |
| `@soil` | Optional soil data (default `NULL`) |

Within the `airGR` package, there are calibration algorithms which allow you to calibrate the hydrological model if discharge data for the reach is available. The calibrated parameters can be passed to `run_GR()` to run the selected model.

```{r run-gr4j, warning=FALSE,fig.height=6,fig.width=8}
# Split the observation period in half for calibration/validation.
# airGR uses row indices; rows 1:(start-1) serve as the warm-up period
# (computed automatically when warmup = NULL).
n_obs   <- nrow(hydro_model@data) - hydro_model@start + 1L
cal_idx <- hydro_model@start:(hydro_model@start + floor(n_obs * 0.5) - 1L)

# Calibrate the model over the calibration period
calib <- calib_GR(hydro_model, run_index = cal_idx)

# Extract the calibrated parameters
param <- calib$ParamFinalR

# Run the model with calibrated parameters
output <- run_GR(hydro_model, param = param, run_index = cal_idx)

# Plot the output against observed discharge
plot(output, Qobs = hydro_model@data$Qmm[cal_idx])
```

