---
title: "Run Hydrological models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run Hydrological models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(aemetools)
```

### Hydrological modelling — Run GR4J model

The workflow follows two steps:

1. **Build the spatial `HydroModel`** using `make_hydro_model()` — this
   assembles the upstream river-channel network, computes the catchment area,
   and stores optional spatial layers (DEM, land cover, soil).
2. **Add airGR inputs** using `make_GR_inputs()` — this attaches the
   meteorological and observed discharge data and constructs the
   `airGR::InputsModel` object.

#### Step 1 — Create the spatial model

```{r make-hydro-model, results='asis', fig.keep='all', warning=FALSE}
data_dir   <- system.file("extdata/hydro/", package = "aemetools")
lake       <- readRDS(file.path(data_dir, "lake.rds"))
reaches    <- readRDS(file.path(data_dir, "reaches.rds"))
catchments <- readRDS(file.path(data_dir, "catchments.rds"))
id         <- 4087861  # nzsegment

hm <- make_hydro_model(id = id, reaches = reaches, lake = lake,
                       catchments = catchments, plot = TRUE)
hm
```

The `HydroModel` S4 object stores the following spatial slots:

| Slot | Description |
|------|-------------|
| `@channels` | Upstream river channels (sf, clipped to exclude lake) |
| `@lake` | Target lake polygon (sf) |
| `@catchments` | Combined upstream catchment polygon with lake removed (sf) |
| `@dem` | Optional DEM raster (default `NULL`) |
| `@land_cover` | Optional land cover data (default `NULL`) |
| `@soil` | Optional soil data (default `NULL`) |
| `@outlet` | Optional outlet points (default `NULL`) |
| `@crs` | Coordinate reference system |
| `@catchment_area` | Catchment area in m² |

#### Step 2 — Attach airGR inputs

```{r make-gr4j-inputs, warning=FALSE}
met      <- readRDS(file.path(data_dir, "met.rds"))
obs_flow <- readRDS(file.path(data_dir, "obs_flow.rds"))

hm <- make_GR_inputs(hm, met = met, obs_flow = obs_flow,
                     FUN_MOD = airGR::RunModel_GR4J)
hm
```

After this step the object also carries the airGR slots:

| Slot | Description |
|------|-------------|
| `@inputs_model` | `airGR` `InputsModel` object |
| `@data` | Combined meteorological and observed flow data |
| `@start` | Row index where the observation period begins |
| `@fun_mod` | `airGR` model function |
| `@fun_mod_name` | Display name of the model function |

#### Calibrate and run

Within the `airGR` package, there are calibration algorithms which allow you to
calibrate the hydrological model if discharge data for the reach is available.
The calibrated parameters can be passed to `run_GR()` to run the selected model.

```{r run-gr4j, warning=FALSE, fig.height=6, fig.width=8}
# Split the observation period in half for calibration/validation.
# airGR uses row indices; rows 1:(start-1) serve as the warm-up period
# (computed automatically when warmup = NULL).
n_obs   <- nrow(hm@data) - hm@start + 1L
cal_idx <- hm@start:(hm@start + floor(n_obs * 0.5) - 1L)

# Calibrate the model over the calibration period
calib <- calib_GR(hm, run_index = cal_idx)

# Extract the calibrated parameters
param <- calib$ParamFinalR

# Run the model with calibrated parameters
output <- run_GR(hm, param = param, run_index = cal_idx)

# Plot the output against observed discharge
plot(output, Qobs = hm@data$Qmm[cal_idx])
```

