---
title: "Screen, Localise, Calibrate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Screen, Localise, Calibrate}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```

Calibrating twenty-odd parameters at once is usually a waste of model runs.
Most of them turn out not to move the variables you care about, and every one
you carry costs ensemble size, run time and - because an unconstrained
parameter still drifts - credibility.

This vignette walks the workflow that avoids that, end to end on a real
monitoring record:

1.  assemble a deliberately **wide** parameter set - heat, mixing, sediment
    heat and sediment nutrient fluxes;
2.  **screen** it with a Morris sensitivity analysis (`pestpp-sen`);
3.  **select** the parameters worth keeping from the screening result;
4.  build a **localizer** from the same numbers, so each variable is only
    updated by the parameters it is actually sensitive to;
5.  **calibrate** the reduced, localised problem with `pestpp-ies`.

Steps 3 and 4 come from one calculation. The screening tells you both *which*
parameters matter and *what they matter to*, and those are exactly the two
things you need.

```{r setup, message = FALSE, warning = FALSE}
library(AEME)
library(aemetools)
```

The PEST++ steps need the solver. Everything below is skipped if it is not
installed - run `install_pest()` once to get it.

```{r have-pest}
run_pest <- have_pest()
run_pest
```

## The lake

Lake Rototoa (LID50270) ships with the package: a 26 m deep dune lake with
GLM-AED configured, roughly 40,000 lake observations spanning 1988-2025, and
daily water level. Its own simulation window is nine years, which is more than
a vignette needs, so we take a single year with a year of spin-up.

```{r lake}
aeme <- readRDS(system.file("extdata/LID50270_Rototoa.rds",
                            package = "aemetools"))
aeme <- AEME::set_time(aeme, start = "2020-07-01", stop = "2021-06-30",
                       spin_up = 365)

vars_sim <- c("HYD_temp", "CHM_oxy", "NIT_amm", "PHS_frp")
```

Four targets: temperature, dissolved oxygen, ammonium and filterable reactive
phosphorus. Oxygen and the nutrients are what give the sediment flux
parameters something to bite on.

## Build and run once

```{r build, eval = run_pest, message = FALSE, results = "hide"}
path <- file.path(tempdir(), "rototoa-screen")
dir.create(path, showWarnings = FALSE, recursive = TRUE)

model_controls <- AEME::set_vars_sim(AEME::get_model_controls(),
                                     vars_sim = vars_sim)
aeme <- AEME::build_aeme(path = path, aeme = aeme, model = "glm_aed",
                         model_controls = model_controls, ext_elev = 5,
                         use_bgc = TRUE)
aeme <- AEME::run_aeme(aeme = aeme, model = "glm_aed", path = path)
```

Both PEST++ steps run their model evaluations in parallel, one process per
core. Take 85% of the machine and leave the rest for everything else:

```{r ncore}
ncore <- max(1L, 2L)
ncore
```

## A wide parameter set

The shipped parameter tables are the starting pool. Hydrodynamic parameters
live in `aeme_parameters`, biogeochemical ones in `aeme_parameters_bgc`, and
the two overlap, so combine and de-duplicate.

```{r pool}
utils::data("aeme_parameters", package = "AEME")
utils::data("aeme_parameters_bgc", package = "AEME")

cols <- c("model", "file", "name", "value", "min", "max", "group", "index")
pool <- unique(rbind(
  aeme_parameters[aeme_parameters$model == "glm_aed", cols],
  aeme_parameters_bgc[aeme_parameters_bgc$model == "glm_aed", cols]
))
pick <- function(nms) as.data.frame(pool[pool$name %in% nms, ])
```

### The index column, and why it is not optional

Sediment parameters are **zoned**. The bed is divided into zones and each
zoned parameter is an array with one value per zone, not a scalar. The `index`
column is how a parameter table says which zone a row refers to.

Getting this wrong does not produce a warning - it produces a model that dies
about 200 days in, because GLM reads a one-element array where it wanted
three.

The catch is that GLM's sediment-heat block and AED's sediment-flux block
declare their zones **separately**, and they need not agree. Read both off the
built configuration rather than assuming:

```{r n-zones, eval = run_pest}
cfg <- AEME::configuration(aeme)$glm_aed

n_glm_zones <- cfg$hydrodynamic$sediment$n_zones
n_aed_zones <- cfg$bgc$aed$aed_sed_const2d$n_zones

c(glm = n_glm_zones, aed = n_aed_zones)
```

Three and two. So the GLM sediment-heat parameters get indices 1:3 and the AED
sediment-flux parameters 1:2. Writing a row for a zone that does not exist is
not fatal here - it is simply ignored, and then shows up in the screening as a
parameter with no effect whatsoever, which is a useful cross-check to have for
free.

```{r per-zone}
per_zone <- function(df, n) {
  do.call(rbind, lapply(seq_len(n), function(i) {
    d <- df
    d$index <- i
    d
  }))
}
```

Independent zone values have a second problem: nothing stops calibration
returning an order that makes no physical sense, such as a zone 1 oxygen flux
larger than zone 2. Shallower zones should not exceed deeper ones, so for the
AED fluxes we use `zone_ratio_param()` instead. Zone 1 (the deepest) is
calibrated directly, and each shallower zone is a ratio between 0.1 and 1 of
the next one down. The ratios are turned back into real per-zone values just
before the model files are written, so the ordering holds by construction and
three weakly constrained values become one value plus two ratios. Sediment
temperature gets the additive version, `zone_offset_param()`: a ratio of
degrees Celsius is arbitrary, and shallower sediment is warmer, so each
shallower zone is the one below plus an offset of 0 to 4 degC.

```{r zone-ratio}
zone_ratios <- function(df, n, start = 0.7) {
  do.call(rbind, lapply(seq_len(nrow(df)), function(i) {
    zone_ratio_param(df[i, ], n_zones = n, start = start)
  }))
}

# Sediment temperature is additive rather than a ratio: shallower zones are
# warmer than deeper ones, so each is the zone below plus an offset
zone_offsets <- function(df, n, upper = 4, start = 0.5) {
  do.call(rbind, lapply(seq_len(nrow(df)), function(i) {
    zone_offset_param(df[i, ], n_zones = n, lower = 0, upper = upper,
                      start = start)
  }))
}

# A ratio/offset row (`<name>_zratio`, `<name>_zoffset`) belongs to the same
# group as its anchor
zgrp <- function(x) sub("_z(ratio|offset)$", "", x)
```

Now the set itself: light and meteorological scaling, the five mixing
coefficients, sediment heat (zoned), the 2-D sediment nutrient fluxes (zoned)
and the bulk sediment flux and half-saturation terms.

```{r param-set, eval = run_pest}
param <- rbind(
  # light, meteorological scaling and mixing
  pick(c("light/Kw", "MET_radswd", "MET_wndspd",
         "mixing/coef_mix_conv", "mixing/coef_mix_hyp",
         "mixing/coef_mix_shear", "mixing/coef_mix_turb",
         "mixing/coef_wind_stir")),
  # sediment heat - zoned by GLM's &sediment block
  zone_offsets(pick("sediment/sed_temp_mean"), n_glm_zones),
  per_zone(pick("sediment/sed_temp_peak_doy"), n_glm_zones),
  # 2-D sediment nutrient/oxygen fluxes - zoned by AED's block
  zone_ratios(pick(c("aed_sed_const2d/fsed_oxy", "aed_sed_const2d/fsed_amm",
                     "aed_sed_const2d/fsed_frp")), n_aed_zones),
  # bulk sediment flux and half-saturation terms
  pick(c("aed_oxygen/fsed_oxy", "aed_oxygen/ksed_oxy",
         "aed_nitrogen/fsed_amm", "aed_nitrogen/ksed_amm",
         "aed_phosphorus/ksed_frp"))
)

# Start the sediment temperature parameters at the observed state: each zone's
# depth band gives an annual mean, amplitude and peak day from the observed
# water temperature. Water temperature is only a proxy for the sediment, so
# these are starting values and the bounds are left as they were.
param <- estimate_zone_start(aeme, param)
zone_start <- attr(param, "zone_estimates")
attr(param, "zone_estimates") <- NULL

param$name_full <- encode_param(param$group, param$name, param$index)

c(parameters = nrow(param), zoned = sum(!is.na(param$index)))
```

`encode_param()` gives each row a unique `name_full` carrying its zone, which
is what everything downstream - the screening output, the localizer, the
results database - keys on:

```{r name-full, eval = run_pest}
head(param$name_full[!is.na(param$index)], 6)
```

## Screen with Morris

`create_sen_control()` defines the *sub-regions* the sensitivity is measured
over - a variable, a set of months and a depth band. This is where the design
work is. Bottom oxygen and bottom ammonium in summer are where a sediment flux
parameter shows up; surface temperature is where the heat parameters do. Ask
the wrong question and everything looks insensitive.

```{r sen-control, eval = run_pest}
z <- AEME::input(aeme)$init_depth

sen_ctrl <- create_sen_control(
  file_dir = file.path(path, "calib_sa"),
  morris_r = 4,
  ncore = ncore,
  vars_sim = list(
    surf_temp = list(var = "HYD_temp", month = c(12, 1, 2),
                     depth_range = c(0, 2)),
    bot_temp  = list(var = "HYD_temp", month = c(12, 1, 2),
                     depth_range = c(z - 4, z)),
    bot_oxy   = list(var = "CHM_oxy",  month = c(12, 1, 2),
                     depth_range = c(z - 4, z)),
    bot_amm   = list(var = "NIT_amm",  month = c(12, 1, 2),
                     depth_range = c(z - 4, z)),
    surf_frp  = list(var = "PHS_frp",  month = c(12, 1, 2),
                     depth_range = c(0, 2))
  )
)
```

Morris costs `morris_r * (k + 1)` model runs, so it scales with the parameter
count rather than exploding with it. The objective is just the mean of the
simulated values in each window: screening asks whether a parameter *moves*
the variable, not whether it improves a fit.

```{r sa-run, eval = run_pest, message = FALSE, results = "hide"}
mean_fit <- function(df) mean(df$model, na.rm = TRUE)
FUN_sa <- stats::setNames(lapply(vars_sim, function(v) mean_fit), vars_sim)

sa_id <- sa_aeme(aeme = aeme, model = "glm_aed", param = param,
                 FUN_list = FUN_sa, path = path, ctrl = sen_ctrl)
sa_id <- unname(sa_id[[1]])
```

`read_sen()` returns the Morris indices per sub-region and parameter.
`mu_star` is the mean absolute elementary effect - overall influence - and
`sigma` measures interaction and non-linearity.

```{r sa-read, eval = run_pest, message = FALSE}
sen <- read_sen(ctrl = sen_ctrl, sim_id = sa_id)
idx <- sen[[sa_id]]$indices
mu <- idx[idx$index_type == "mu_star", ]

plot_sen(sen)
```

## Select what to keep

`mu_star` is in the units of each sub-region's own objective, so the values
are not comparable across sub-regions. Scale each to its own maximum first,
then keep any parameter that reaches `thr` for at least one sub-region.

```{r select, eval = run_pest}
thr <- 0.10

mu$rel <- stats::ave(mu$value, mu$variable, FUN = function(v) {
  if (max(v, na.rm = TRUE) > 0) v / max(v, na.rm = TRUE) else v * 0
})

# Which AEME variable does each sub-region belong to?
sub_var <- vapply(sen_ctrl$vars_sim, function(x) x$var, character(1))
mu$var_aeme <- sub_var[mu$variable]

influential <- mu[mu$rel >= thr, ]
keep <- unique(influential$parameter)

# An anchor and its ratios only make sense together: keep the whole group if
# any member is influential.
zr_grp <- unique(zgrp(param$name[grepl("_z(ratio|offset)$", param$name)]))
in_zr <- zgrp(param$name) %in% zr_grp
keep_zr <- unique(zgrp(param$name[param$name_full %in% keep]))
keep <- unique(c(keep, param$name_full[in_zr & zgrp(param$name) %in% keep_zr]))

param_keep <- param[param$name_full %in% keep, ]
c(kept = nrow(param_keep), dropped = nrow(param) - nrow(param_keep))
```

A parameter that reaches 10% of the leading effect for nothing at all is not
worth an ensemble slot. What gets dropped is worth reading rather than taking
on trust - here it is several mixing coefficients and the sediment
peak-day terms:

```{r dropped, eval = run_pest}
setdiff(param$name_full, keep)
```

Note that the zones separate: a zoned parameter can matter in one sediment zone
and not another. That is the cross-check mentioned earlier - had the GLM
sediment parameters been given four zones instead of three, every `[4]` row
would appear in this list, because a parameter written to a zone the model
does not have cannot move anything.

Dropping individual zones is not safe in general. GLM's `&sediment` block does
not tolerate a partial set: writing `sed_temp_mean` for one zone out of three
produces a short array and the model dies a couple of hundred days in. The
zone-ratio parameters are stricter still, because a ratio is meaningless
without its anchor, so the selection above keeps each anchor and its ratios as
one group.

So the rule to carry away is that a zoned parameter is kept or dropped as a
whole set, on the strength of its most influential zone, unless you know the
block tolerates otherwise. `vignette("staged-calibration")` does that, and
shows what it costs.

## Build the localizer

The same table says *what* each parameter influences. A localizer is that
statement in the form `pestpp-ies` wants: a matrix of observation groups by
parameters, where a zero severs the link so an oxygen residual can never move
a parameter that only affects temperature.

```{r localizer, eval = run_pest}
# A zone-ratio group has no sensitivity of its own for the members that were
# re-added to complete it, and pestpp-ies treats an unlinked parameter as
# fixed, so link the whole group to the variables its influential member moves.
pvm_list <- lapply(split(influential$parameter, influential$var_aeme), function(z) {
  z <- unique(z)
  gz <- intersect(zgrp(param$name[param$name_full %in% z]), zr_grp)
  param_keep$name_full[param_keep$name_full %in% z |
                         zgrp(param_keep$name) %in% gz]
})
pvm <- as_param_var_matrix(pvm_list, param = param_keep, vars_sim = vars_sim)

colSums(pvm[, vars_sim, drop = FALSE])
```

Without this, every residual updates every parameter, and with a modest
ensemble that is how spurious correlations creep into the posterior.

## Calibrate

Pass the reduced parameter set and the localizer. `calib_aeme()` writes the
localizer, the template, the instruction file and the control file, checks the
forward run once, then hands over to the solver.

```{r calibrate, eval = run_pest, message = FALSE, results = "hide"}
cal_ctrl <- create_pest_control(
  exe = "pestpp-ies",
  noptmax = 2,
  ies_num_reals = 20,
  ncore = ncore,
  file_dir = file.path(path, "calib_sa"),
  pest_dir = "pest_calib",
  localizer = pvm,
  pestpp_options = list(ies_lambda_mults = 1, lambda_scale_fac = 1)
)

FUN_cal <- stats::setNames(lapply(vars_sim, function(v) kge_loss), vars_sim)

cal_id <- calib_aeme(aeme = aeme, path = path, param = param_keep,
                     model = "glm_aed", vars_sim = vars_sim,
                     FUN_list = FUN_cal, weights = set_weights(vars_sim),
                     ctrl = cal_ctrl)
cal_id <- unname(cal_id[[1]])
```

The run history comes back through the ordinary readers, so everything that
works for the built-in methods works here too.

```{r results, eval = run_pest}
calib <- read_calib(ctrl = cal_ctrl, sim_id = cal_id)

# Objective function per iteration.
read_pest_phi(calib)[, c("iteration", "total_runs", "mean", "min")]

# Best parameter set.
best <- get_best_params(calib = calib)
head(best[, c("name", "index", "value", "min", "max")])
```

Per-group phi is the thing to look at next: it shows whether one variable is
dominating the objective, which is what `pest_adjust_weights()` is for.

```{r phi-group, eval = run_pest}
pg <- read_pest_phi_group(calib)
aggregate(phi ~ obgnme, data = pg[pg$iteration == max(pg$iteration), ],
          FUN = mean)
```

And `pest_prior_data_conflict()` flags observations the prior ensemble cannot
reproduce at all - a model-structure signal rather than a parameter one, and
worth checking before you believe any of the above.

## Where to go next

-   `get_calib_periods()` splits the observation record into calibration and
    validation periods, or takes the whole record with `split = FALSE`; see
    `vignette("calibrate-aeme")` for the split-sample argument.
-   `pest_adjust_weights()` rebalances the observation groups after a first
    run so no one variable dominates.
-   Raise `morris_r` for a screening you intend to act on. Four trajectories
    is enough to demonstrate the workflow; it is thin for a real decision.
