---
title: "Sensitivity Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrate AEME}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
library(printr)
unlink("sa-test", recursive = TRUE, force = TRUE)
```

## Setup

First, we will load the `AEME` and `aemetools` package:

```{r setup, message = FALSE, warning=FALSE}
library(AEME)
library(aemetools)
```

Create a folder for running the example calibration setup.

```{r create-folder}

tmpdir <- "sa-test"
dir.create(tmpdir, showWarnings = FALSE)
aeme_dir <- system.file("extdata/lake/", package = "AEME")
# Copy files from package into tempdir
file.copy(aeme_dir, tmpdir, recursive = TRUE)
path <- file.path(tmpdir, "lake")

list.files(path, recursive = TRUE)
```

## Build AEME ensemble

Using the `AEME` functions, we will build the AEME model setup. For this
example, we will use the `glm_aed` model. The `build_ensemble` function will

```{r build-aeme-example, warning=FALSE, message=FALSE}

aeme <- yaml_to_aeme(path = path, "aeme.yaml")
model_controls <- AEME::get_model_controls()
inf_factor = c("dy_cd" = 1, "glm_aed" = 1, "gotm_wet" = 1)
outf_factor = c("dy_cd" = 1, "glm_aed" = 1, "gotm_wet" = 1)
model <- c("gotm_wet")
aeme <- build_ensemble(path = path, aeme = aeme,
                       model = model, model_controls = model_controls,
                       inf_factor = inf_factor, ext_elev = 5,
                       use_bgc = TRUE)

```

## Description of Sensitivity Analysis method

The sensitivity analysis method used here is based on the Sobol method 
and uses the `sensobol` package.

This package provides several functions to conduct variance-based uncertainty 
and sensitivity analysis, from the estimation of sensitivity indices to the 
visual representation of the results. It implements several state-of-the-art 
first and total-order estimators and allows the computation of up to 
fourth-order effects, as well as of the approximation error, in a swift and 
user-friendly way.

For more information on the method, see the [sensobol package
vignette](https://cran.r-project.org/web/packages/sensobol/vignettes/sensobol.html).


## Load parameters to be used for the sensitivity analysis

Parameters are loaded from the `aemetools` package within the `aeme_parameters`
dataframe. The parameters are stored in a data frame with the following columns:

-   `model`: The model name

-   `file`: The file name of the model parameter file

-   `name`: The parameter name

-   `value`: The parameter value

-   `min`: The minimum value of the parameter

-   `max`: The maximum value of the parameter

Parameters to be used for the calibration. (man)

```{r}
#| label: load-parameters
#| tbl-captions: "Parameters to be used for the calibration."
utils::data("aeme_parameters", package = "aemetools")
param <- aeme_parameters |>
  dplyr::filter(file != "wdr")
param
```

## Sensitivity analysis setup

### Define fitness function

First, we will define a function for the sensitivity analysis function to use to
calculate the sensitivity of the model. This function takes a dataframe as an
argument. The dataframe contains the observed data (`obs`) and the modelled data
(`model`). The function should return a single value.

Here we use the model mean.

```{r define-fitness-function}
# Function to calculate mean model output
fit <- function(df) {
  mean(df$model)
}
```

Different functions can be applied to different variables. For example, we can
use the mean for water temperature and median for chloophyll-a.

```{r define-fitness-function-2}
# Function to calculate median model output
fit2 <- function(df) {
  median(df$model)
}
```

Then these would be combined into a named list of functions which will be passed
to the `sa_aeme` function. They are named according to the target variable.

```{r define-function-list}

# Create list of functions
FUN_list <- list(HYD_temp = fit, PHY_tchla = fit2)

```

### Define control parameters

Next, we will define the control parameters for the sensitivity analysis. The
control parameters are generated using `create_control` and are then passed to 
the `sa_aeme` function. The control parameters for the sensitivity analysis are 
as follows:

```{r create-control-args, tidy=FALSE, printr.help.sections=c('arguments'), comment=''}
?create_control
```

Here is an example for examining surface temperature (surf_temp) in the months
December to February, bottom temperature (bot_temp), (10 - 13 m) and also total
chlorophyll-a (PHY_tchla) at the surface (0 - 2 m) during the summer period.

```{r define-control-parameters}
ctrl <- create_control(method = "sa", N = 2^4, ncore = 2, na_value = 999,
                       parallel = TRUE, file_type = "db", 
                       file_name = "file_type",
                       vars_sim = list(
                         surf_temp = list(var = "HYD_temp",
                                          month = c(12, 1:2),
                                          depth_range = c(0, 2) 
                         ),
                         bot_temp = list(var = "HYD_temp",
                                         month = c(12, 1:2),
                                         depth_range = c(10, 13)
                         ),
                         surf_chla = list(var = "PHY_tchla",
                                          month = c(12, 1:2),
                                          depth_range = c(0, 2)
                         )
                       )
)
```

## Run sensitivity analysis

Once we have defined the fitness function, control parameters and variables, we
can run the sensitivity analysis. The `sa_aeme` function takes the following
arguments:

```{r sa-aeme-args, tidy=FALSE, printr.help.sections=c('arguments'), comment=''}
?sa_aeme
```


The `sa_aeme` function writes the results to the file specified.  The 
`sa_aeme` function returns the `sim_id` of the run.

```{r run-sensitivity-analysis, cache=FALSE}
# Run sensitivity analysis AEME model
sim_id <- sa_aeme(aeme = aeme, path = path, param = param,
                  model = model, ctrl = ctrl, model_controls = model_controls,
                  FUN_list = FUN_list)

```

## Reading sensitivity analysis results

The sensitivity results can be read in using the `read_sa` function. This
function takes the following arguments:

-   `ctrl`: The control parameters used for the sensitivity analysis.
-   `model`: The model used for the sensitivity analysis.
-   `path`: The path to the directory where the model is configuration is.

```{r read-sensitivity-analysis, warning=FALSE}
# Read in sensitivity analysis results
sa_res <- read_sa(ctrl = ctrl, sim_id = sim_id, R = 10^3)
names(sa_res)
```

The `read_sa` function returns a list for each simulation id provided. This list
contains the following elements:

-   `df`: dataframe of the sensitivity analysis results. The dataframe contains
the model, generation, index (model run), parameter name, parameter value,
fitness value and the median fitness value for each generation.

```{r print-sa-df}
head(sa_res[[1]]$df)
```

-   `sobol_indices`: list of the Sobol indices for each variable an it's 
senstivity to the parameters.

```{r print-sa-sobol}
sa_res[[1]]$sobol_indices
```

-   `sobol_dummy`: list of the Sobol indices for the dummy parameter.

```{r print-sa-sobol-dummy}
sa_res[[1]]$sobol_dummy

```

## Visualising sensitivity analysis results

The sensitivity analysis results can be visualised in different ways using the
functions: `plot_uncertainty`, `plot_scatter` and `plot_multiscatter`. These 
plots are based on the output plots from the `sensobol` package. 

These functions take the following argument:

-   `sa_res`: The sensitivity analysis results returned from the `read_sa`
function.

### Uncertainty plot

The `plot_uncertainty` function plots the distribution of the model output for
each variable.

```{r plot-sa-uncertainty, fig.width=8, fig.height=6}
# Plot sensitivity analysis results
plot_uncertainty(sa_res)
```

### Scatter plot

The `plot_scatter` function plots the model output against the parameter value
for each variable. This is useful for identifying relationships between the
model output and the parameter value. For example, the plot below shows that
there is a relationship between the model surface temperature (surf_temp_) and 
the parameter value of the scaling factor for shortwave radiation (MET_radswd), 
and also for surface chlorophyll-a (surf_chla) and the light extinction
coefficient (light.Kw). When there is a low parameter value for Kw, the model
chlorophyll-a is higher.

```{r plot-sa-scatter, fig.width=8, fig.height=6}
plot_scatter(sa_res)
```

### Multi-scatter plot

The `plot_multiscatter` function plots the parameters against each other for
each variable. The parameter on top is the x-axis and the parameter below is 
the y-axis. This is useful for identifying relationships between the
parameters and response variable.

```{r plot-sa-multiscatter, fig.width=8, fig.height=6}
pl <- plot_multiscatter(sa_res)

pl[[1]][1]

pl[[1]][2]

pl[[1]][3]

```
