---
title: "Staged Calibration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Staged Calibration}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE)
```

Calibrating a lake's physics and biogeochemistry at once asks the solver to
decide whether too little bottom oxygen is a sediment flux problem, a mixing
problem or a water balance problem. It cannot tell, so it compensates: a
plausible fit built out of offsetting errors.

Staging removes the ambiguity by exploiting the asymmetry. Physics drives
biogeochemistry - stratification controls oxygen and nutrient recycling - but
biogeochemistry barely feeds back. So calibrate the water balance first, then
the thermal structure, freeze both, and only then let the biogeochemical
parameters move.

Each stage is screened, localised and solved the same way, as in
`vignette("screen-then-calibrate")`:

| stage | targets | parameters |
|---|---|---|
| 1 | water level | inflow, outflow, rainfall, seepage |
| 2 | temperature + level | light, met scaling, mixing, sediment heat |
| 3 | oxygen, ammonium, phosphorus | sediment fluxes, N/P kinetics, phytoplankton |

The whole workflow is also available as a plain script:

```{r script-path, eval = FALSE}
file.edit(system.file("scripts/staged-calibration.R", package = "aemetools"))
```

```{r setup, message = FALSE, warning = FALSE}
library(AEME)
library(aemetools)
run_pest <- have_pest()
```

```{r config, eval = run_pest}
ROOT <- file.path(tempdir(), "staged")
MODEL <- "glm_aed"
NCORE <- max(1L, floor(0.85 * parallel::detectCores()))   # 85% of the box
THR <- 0.10

aeme0 <- readRDS(system.file("extdata/LID50270_Rototoa.rds",
                             package = "aemetools"))
aeme0 <- AEME::set_time(aeme0, start = "2020-07-01", stop = "2021-06-30",
                        spin_up = 365)
NCORE
```

## Two things that will bite you

### Zoned parameters are arrays, not scalars

Sediment parameters have one value per bed zone. The `index` column says which
zone a row refers to, and a parameter written with fewer rows than the model
has zones becomes a short array - GLM then dies a couple of hundred days in,
with no useful message.

Worse, GLM's `&sediment` block and AED's sediment block declare their zone
counts **separately**, and here they disagree:

```{r zones-build, eval = run_pest, message = FALSE, results = "hide"}
probe <- AEME::build_aeme(path = file.path(ROOT, "probe"), aeme = aeme0,
                          model = MODEL, use_bgc = TRUE, ext_elev = 5,
                          model_controls = AEME::set_vars_sim(
                            AEME::get_model_controls(),
                            vars_sim = c("HYD_temp", "CHM_oxy")))
cfg <- AEME::configuration(probe)$glm_aed
```

```{r zones, eval = run_pest}
c(glm = cfg$hydrodynamic$sediment$n_zones,
  aed = cfg$bgc$aed$aed_sed_const2d$n_zones)
```

The consequence for screening is not obvious: **a zoned parameter has to be
kept or dropped as a whole set.** If the screen finds zone 2 influential and
zones 1 and 3 not, keeping only zone 2 writes a one-element array and the
calibration fails before it starts. The `run_stage()` helper below re-adds the
missing zones.

For the AED sediment fluxes there is a second problem: three independent
zone values can come back in an order that makes no physical sense, such as a
zone 1 oxygen flux larger than zone 2. Stage 3 therefore parameterises them
with `zone_ratio_param()`: zone 1 (the deepest) is calibrated directly, and
each shallower zone is a ratio (0.1 to 1) of the next one down. Ratios are
expanded back into real per-zone values immediately before the model files are
written, so a shallower zone can never exceed a deeper one. Sediment
temperature in stage 2 uses the additive version, `zone_offset_param()`, since
a ratio of degrees Celsius is arbitrary and shallower sediment is warmer, not
cooler: each shallower zone is the one below plus an offset of 0 to 4 degC.

### Every step needs its own build

PEST staging and `input_model_parameters()` both rewrite the lake directory in
place. A stage - or even a calibration following a screen - run in a directory
a previous step has touched starts from a configuration that is no longer the
one that was built, and every model run fails.

```{r fresh, eval = run_pest}
fresh_build <- function(tag, vars_sim, use_bgc) {
  p <- file.path(ROOT, tag)
  unlink(p, recursive = TRUE)
  dir.create(p, recursive = TRUE, showWarnings = FALSE)
  mc <- AEME::set_vars_sim(AEME::get_model_controls(), vars_sim = vars_sim)
  a <- AEME::build_aeme(path = p, aeme = aeme0, model = MODEL,
                        model_controls = mc, ext_elev = 5, use_bgc = use_bgc)
  list(aeme = AEME::run_aeme(aeme = a, model = MODEL, path = p), path = p)
}
```

## Parameter helpers

```{r helpers, eval = run_pest}
utils::data("aeme_parameters", package = "AEME")
utils::data("aeme_parameters_bgc", package = "AEME")

PCOLS <- c("model", "file", "name", "value", "min", "max", "group", "index")
POOL <- unique(rbind(
  aeme_parameters[aeme_parameters$model == MODEL, PCOLS],
  aeme_parameters_bgc[aeme_parameters_bgc$model == MODEL, PCOLS]))

shipped <- function(nms) as.data.frame(POOL[POOL$name %in% nms, ])

prow <- function(file, name, value, min = NULL, max = NULL, fac = 0.5,
                 group = NA_character_, index = NA_real_) {
  if (is.null(min)) min <- value - abs(value) * fac
  if (is.null(max)) max <- value + abs(value) * fac
  data.frame(model = MODEL, file = file, name = name, value = value,
             min = min, max = max, group = group, index = index,
             stringsAsFactors = FALSE)
}

per_zone <- function(df, n) do.call(rbind, lapply(seq_len(n), function(i) {
  d <- df; d$index <- i; d }))
per_phyto <- function(df, groups) do.call(rbind, lapply(groups, function(g) {
  d <- df; d$group <- g; d }))

# A zone-ratio parameter (`<name>_zratio`, see `zone_ratio_param()`) belongs to
# the same group as its anchor: screening keeps or drops them together.
zgrp <- function(x) sub("_z(ratio|offset)$", "", x)

add_name_full <- function(p) { p$name_full <- encode_param(p$group, p$name, p$index); p }

# value == min == max freezes a parameter. sa_aeme() and the built-in
# calib engines write it to the model and drop it; a PEST run keeps it in
# the .pst as `partrans = fixed`, so it stays visible in the parameter map,
# the ensembles and the sensitivity output.
freeze <- function(p) { p$min <- p$value; p$max <- p$value; p }
carry <- function(res, except = character()) {
  freeze(res$best[res$best$name_full %in% res$kept$name_full &
                    !(res$best$name %in% except), PCOLS])
}
```

## One stage

Screen with Morris, keep what moves the targets, turn the same numbers into a
localizer, calibrate. `screen = FALSE` skips the screen, which is right for a
handful of parameters and *required* for a water-level-only stage - the
sensitivity path has no gridded variable to index.

```{r run-stage, eval = run_pest}
run_stage <- function(stage, aeme, path, adjustable, frozen, targets,
                      sub_regions, screen = TRUE, rebuild = NULL,
                      prior_from = NULL) {

  adj <- add_name_full(adjustable)
  fdir <- file.path(path, "calib_sa")
  keep_adj <- adj; pvm <- NULL; sen <- NULL

  if (screen) {
    sen_ctrl <- create_sen_control(file_dir = fdir, morris_r = 4,
                                   ncore = NCORE, vars_sim = sub_regions)
    mean_fit <- function(df) mean(df$model, na.rm = TRUE)
    sa_id <- unname(sa_aeme(
      aeme = aeme, model = MODEL,
      param = add_name_full(rbind(adj[, PCOLS], frozen[, PCOLS])),
      FUN_list = stats::setNames(lapply(targets, function(v) mean_fit), targets),
      path = path, ctrl = sen_ctrl)[[1]])

    sen <- read_sen(ctrl = sen_ctrl, sim_id = sa_id)
    mu <- sen[[sa_id]]$indices
    mu <- mu[mu$index_type == "mu_star", ]
    mu$rel <- stats::ave(mu$value, mu$variable, FUN = function(v)
      if (max(v, na.rm = TRUE) > 0) v / max(v, na.rm = TRUE) else v * 0)
    sub_var <- vapply(sub_regions, function(x) x$var, character(1))
    mu$var_aeme <- sub_var[mu$variable]
    infl <- mu[mu$rel >= THR, ]

    # Complete zone sets, as explained above.
    keep_names <- unique(zgrp(adj$name[adj$name_full %in% unique(infl$parameter)]))
    keep_adj <- adj[zgrp(adj$name) %in% keep_names, ]

    # Link by name too: a zone re-added to complete an array has no
    # sensitivity of its own, and pestpp-ies treats an unlinked parameter as
    # fixed.
    pvm_list <- lapply(split(infl$parameter, infl$var_aeme), function(z) {
      nms <- unique(zgrp(adj$name[adj$name_full %in% unique(z)]))
      keep_adj$name_full[zgrp(keep_adj$name) %in% nms]
    })
    pvm <- as_param_var_matrix(pvm_list[lengths(pvm_list) > 0],
                               param = keep_adj, vars_sim = targets)
  }

  if (!is.null(rebuild)) {
    b <- rebuild(); aeme <- b$aeme; path <- b$path
    fdir <- file.path(path, "calib_sa")
  }

  prior_en <- if (is.null(prior_from)) NULL else
    read_calib(ctrl = prior_from$cal_ctrl, sim_id = prior_from$cal_id)

  cal_ctrl <- create_pest_control(
    exe = "pestpp-ies", noptmax = 2, ies_num_reals = 20, ncore = NCORE,
    file_dir = fdir, pest_dir = paste0("pest_s", stage), localizer = pvm,
    prior_par_ensemble = prior_en,
    pestpp_options = list(ies_lambda_mults = 1, lambda_scale_fac = 1))

  cal_id <- unname(calib_aeme(
    aeme = aeme, path = path,
    param = add_name_full(rbind(keep_adj[, PCOLS], frozen[, PCOLS])),
    model = MODEL, vars_sim = targets,
    FUN_list = stats::setNames(lapply(targets, function(v) kge_loss), targets),
    weights = set_weights(targets), ctrl = cal_ctrl)[[1]])

  calib <- read_calib(ctrl = cal_ctrl, sim_id = cal_id)
  list(stage = stage, kept = keep_adj, pvm = pvm, sen = sen,
       cal_ctrl = cal_ctrl, cal_id = cal_id,
       phi = read_pest_phi(calib),
       best = add_name_full(get_best_params(calib = calib)[, PCOLS]))
}
```

## Stage 1 - water balance

Four parameters against water level. Nothing to screen out.

```{r stage1, eval = run_pest, message = FALSE, results = "hide"}
b1 <- fresh_build("s1", "LKE_lvlwtr", use_bgc = FALSE)
cfg <- AEME::configuration(b1$aeme)$glm_aed
sed <- cfg$hydrodynamic$sediment
n_glm_zones <- sed$n_zones
z <- AEME::input(b1$aeme)$init_depth

s1 <- run_stage(1, b1$aeme, b1$path,
                adjustable = rbind(
                  shipped(c("inflow", "outflow")),
                  prow("met", "MET_pprain", 1, min = 0.7, max = 1.3),
                  prow("glm4.nml", "outflow/seepage_rate", 5e-4,
                       min = 0, max = 2e-3)),
                frozen = POOL[0, PCOLS], targets = "LKE_lvlwtr",
                sub_regions = NULL, screen = FALSE)
```

```{r stage1-phi, eval = run_pest}
s1$phi[, c("iteration", "total_runs", "min")]
```

## Stage 2 - thermal structure

Light, meteorological scaling, the mixing coefficients, and sediment heat
across all three GLM zones. Stage 1's winners are frozen.

The sediment temperature parameters start from the observations rather than
the shipped defaults. `estimate_zone_start()` turns each zone into a depth
band (from GLM's `zone_heights` and the lake depth) and fits an annual cycle
to the observed water temperature in it, giving a mean, amplitude and peak
day per zone. The zone 1 mean becomes the anchor, and the differences between
neighbouring zones become the offsets, clipped to their bounds. Water
temperature is only a proxy, since real sediment is damped and lags it, so
these are starting values and the bounds still do the work.

`prior_from = s1` seeds this stage's prior ensemble from stage 1's
*posterior* rather than from independent uniforms on the bounds
(`create_pest_control(prior_par_ensemble = ...)`, given a `read_calib`
object). An **adjustable** parameter that stage 1 also calibrated comes in at
its stage-1 posterior spread and correlations; a parameter new to this stage
is drawn from its own bounds; a frozen parameter is written at its fixed
value. Here stage 2's adjustable set is disjoint from stage 1's, so the gain
shows at stage 3, where `light/Kw` - re-opened after being fixed by stage 2 -
starts from stage 2's posterior for it rather than from a fresh guess.

```{r stage2, eval = run_pest, message = FALSE, results = "hide"}
vars_phys <- c("HYD_temp", "LKE_lvlwtr")
b2 <- fresh_build("s2", vars_phys, use_bgc = FALSE)

s2_adj <- rbind(
  shipped(c("light/Kw", "MET_radswd", "MET_wndspd",
            "mixing/coef_mix_conv", "mixing/coef_mix_hyp",
            "mixing/coef_mix_shear", "mixing/coef_mix_turb",
            "mixing/coef_wind_stir")),
  prow("glm4.nml", "mixing/coef_mix_KH", cfg$hydrodynamic$mixing$coef_mix_KH),
  prow("met", "MET_tmpair", 1, min = 0.9, max = 1.1),
  # Zone 1 (deepest) temperature + a 0-4 degC warming for each shallower zone
  zone_offset_param(prow("glm4.nml", "sediment/sed_temp_mean",
                         mean(sed$sed_temp_mean), fac = 0.4, index = 1),
                    n_glm_zones, lower = 0, upper = 4, start = 0.5),
  per_zone(prow("glm4.nml", "sediment/sed_temp_amplitude",
                mean(sed$sed_temp_amplitude), min = 0.1, max = 6), n_glm_zones),
  per_zone(prow("glm4.nml", "sediment/sed_heat_Ksoil",
                mean(sed$sed_heat_Ksoil), fac = 0.6), n_glm_zones),
  per_zone(prow("glm4.nml", "sediment/sed_temp_depth",
                mean(sed$sed_temp_depth), fac = 0.6), n_glm_zones))

# Start the sediment temperature parameters at the observed state rather than
# at the shipped defaults: each zone's depth band gives an annual mean,
# amplitude and peak day from the observed water temperature.
s2_adj <- estimate_zone_start(b2$aeme, s2_adj)
s2_start <- attr(s2_adj, "zone_estimates")
attr(s2_adj, "zone_estimates") <- NULL

s2 <- run_stage(
  2, b2$aeme, b2$path, s2_adj, carry(s1), targets = vars_phys,
  sub_regions = list(
    surf_temp = list(var = "HYD_temp", month = c(12, 1, 2), depth_range = c(0, 2)),
    bot_temp  = list(var = "HYD_temp", month = c(12, 1, 2), depth_range = c(z - 4, z)),
    win_temp  = list(var = "HYD_temp", month = c(6, 7, 8), depth_range = c(0, z)),
    lake_lvl  = list(var = "LKE_lvlwtr", month = 1:12, depth_range = c(0, 1))),
  rebuild = function() fresh_build("s2cal", vars_phys, use_bgc = FALSE),
  prior_from = s1)
```

```{r stage2-out, eval = run_pest}
c(pool = 22, kept = nrow(s2$kept))
s2$phi[, c("iteration", "total_runs", "min")]
colSums(s2$pvm[, vars_phys, drop = FALSE])
```

## Stage 3 - biogeochemistry

Now `use_bgc = TRUE`. Water level is deliberately **not** a target here:
combining it with biogeochemistry returns empty variable indices, and the
water balance is fixed by stage 1 anyway.

```{r stage3-setup, eval = run_pest, message = FALSE, results = "hide"}
vars_bgc <- c("CHM_oxy", "NIT_amm", "PHS_frp")
b3 <- fresh_build("s3", vars_bgc, use_bgc = TRUE)
cfg3 <- AEME::configuration(b3$aeme)$glm_aed
aed <- cfg3$bgc$aed
n_aed_zones <- aed$aed_sed_const2d$n_zones
phyto <- as.data.frame(cfg3$bgc$aed_phyto_pars)
groups <- c("cyano", "green", "diatom")     # the_phytos = 2, 3, 7

pv <- function(nm, g) as.numeric(phyto[phyto$p_name == nm, g])
av <- function(mod, nm) as.numeric(aed[[mod]][[nm]][1])
```

Phytoplankton parameters live in `aed_phyto_pars.csv`, which is keyed the
other way round from a namelist: the file's **rows** are parameters and its
**columns** are groups. So `name` selects the row and `group` selects the
column - which is what the `group` column is for, exactly as `index` selects a
sediment zone.

```{r stage3-pool, eval = run_pest, message = FALSE, results = "hide"}
s3_adj <- rbind(
  # Zone 1 value + a 0.1-1 ratio for each shallower zone
  zone_ratio_param(prow("aed.nml", "aed_sed_const2d/fsed_oxy",
                        av("aed_sed_const2d", "fsed_oxy"), fac = 0.5,
                        index = 1), n_aed_zones, start = 0.7),
  zone_ratio_param(prow("aed.nml", "aed_sed_const2d/fsed_amm",
                        av("aed_sed_const2d", "fsed_amm"), fac = 0.7,
                        index = 1), n_aed_zones, start = 0.7),
  zone_ratio_param(prow("aed.nml", "aed_sed_const2d/fsed_frp",
                        av("aed_sed_const2d", "fsed_frp"), fac = 0.7,
                        index = 1), n_aed_zones, start = 0.7),
  prow("aed.nml", "aed_oxygen/fsed_oxy", av("aed_oxygen", "fsed_oxy"), fac = 0.5),
  prow("aed.nml", "aed_oxygen/ksed_oxy", av("aed_oxygen", "ksed_oxy"), fac = 0.5),
  prow("aed.nml", "aed_oxygen/theta_sed_oxy",
       av("aed_oxygen", "theta_sed_oxy"), min = 1.02, max = 1.12),
  prow("aed.nml", "aed_nitrogen/rnitrif", av("aed_nitrogen", "rnitrif"), fac = 0.6),
  prow("aed.nml", "aed_nitrogen/knitrif", av("aed_nitrogen", "knitrif"), fac = 0.5),
  prow("aed.nml", "aed_nitrogen/rdenit", av("aed_nitrogen", "rdenit"), fac = 0.6),
  prow("aed.nml", "aed_nitrogen/kdenit", av("aed_nitrogen", "kdenit"), fac = 0.5),
  prow("aed.nml", "aed_nitrogen/ksed_amm", av("aed_nitrogen", "ksed_amm"), fac = 0.5),
  prow("aed.nml", "aed_phosphorus/ksed_frp",
       av("aed_phosphorus", "ksed_frp"), fac = 0.5),
  prow("aed.nml", "aed_phosphorus/theta_sed_frp",
       av("aed_phosphorus", "theta_sed_frp"), min = 1.02, max = 1.12),
  prow("aed.nml", "aed_organic_matter/rdom_minerl",
       av("aed_organic_matter", "rdom_minerl"), fac = 0.7),
  prow("aed.nml", "aed_organic_matter/kpom_hydrol",
       av("aed_organic_matter", "kpom_hydrol"), fac = 0.5),
  prow("aed.nml", "aed_organic_matter/kdom_minerl",
       av("aed_organic_matter", "kdom_minerl"), fac = 0.5),
  prow("aed.nml", "aed_organic_matter/fsed_doc",
       av("aed_organic_matter", "fsed_doc"), fac = 0.7),
  per_phyto(prow("aed_phyto_pars.csv", "R_growth", pv("R_growth", "cyano"),
                 fac = 0.5), groups),
  per_phyto(prow("aed_phyto_pars.csv", "I_K", pv("I_K", "cyano"), fac = 0.6), groups),
  per_phyto(prow("aed_phyto_pars.csv", "K_N", pv("K_N", "cyano"), fac = 0.6), groups),
  per_phyto(prow("aed_phyto_pars.csv", "K_P", pv("K_P", "cyano"), fac = 0.6), groups),
  per_phyto(prow("aed_phyto_pars.csv", "R_resp", pv("R_resp", "cyano"),
                 fac = 0.6), groups),
  # Light attenuation re-opened, bounded by the stage-2 posterior: it is the
  # one place the thermal and biogeochemical calibrations genuinely meet.
  s2$best[s2$best$name == "light/Kw", PCOLS])

s3 <- run_stage(
  3, b3$aeme, b3$path, s3_adj,
  rbind(carry(s1), carry(s2, except = "light/Kw")), targets = vars_bgc,
  sub_regions = list(
    bot_oxy  = list(var = "CHM_oxy", month = c(12, 1, 2), depth_range = c(z - 4, z)),
    surf_oxy = list(var = "CHM_oxy", month = c(12, 1, 2), depth_range = c(0, 2)),
    bot_amm  = list(var = "NIT_amm", month = c(12, 1, 2), depth_range = c(z - 4, z)),
    surf_frp = list(var = "PHS_frp", month = c(12, 1, 2), depth_range = c(0, 2))),
  rebuild = function() fresh_build("s3cal", vars_bgc, use_bgc = TRUE),
  prior_from = s2)
```

```{r stage3-out, eval = run_pest}
c(pool = nrow(s3_adj), kept = nrow(s3$kept))
s3$kept[, c("file", "name", "group", "index")]
```

Every phytoplankton parameter was dropped, and that is the correct answer
rather than a failure: with oxygen, ammonium and phosphorus as the only
targets there is nothing in the objective that growth rate or light
saturation can move. **Phytoplankton parameters need a chlorophyll target to
be identifiable.** `PHY_tchla` cannot currently be used as a sensitivity
sub-region here - it trims to fewer timesteps than requested and the screen
fails - so the honest outcome is to leave those parameters at their defaults
and say so.

## Results

```{r results, eval = run_pest}
data.frame(
  stage = c(1, 2, 3),
  targets = c("water level", "temperature + level", "oxygen, ammonium, FRP"),
  kept = c(nrow(s1$kept), nrow(s2$kept), nrow(s3$kept)),
  phi_start = c(s1$phi$min[1], s2$phi$min[1], s3$phi$min[1]),
  phi_end = c(utils::tail(s1$phi$min, 1), utils::tail(s2$phi$min, 1),
              utils::tail(s3$phi$min, 1)))
```

Stages 1 and 2 improve substantially. Stage 3 does not move at all - the
retained sediment fluxes cannot shift the objective within two iterations of a
twenty-member ensemble. Read that as a result, not a formality: either the
ensemble and iteration budget are too small, or the model's oxygen and
nutrient errors are structural rather than parametric. Checking
`pest_prior_data_conflict()` distinguishes the two, and raising `noptmax` and
`ies_num_reals` is the cheap experiment.

## Posterior predictive ensemble

`pestpp-ies` is an iterative ensemble smoother. Its real product is not the
single best fit but the **posterior ensemble** - one parameter vector per
realisation - and the honest way to report a calibration is to push that
ensemble back through the model and look at the spread.

`pest_posterior_params()` reads the final `<case>.<n>.par.csv` (the posterior;
`iteration = 0` gives the prior) and returns one runnable `param` dataframe per
realisation. Parameters that were frozen in earlier stages are absent from the
solver's ensemble, so they are carried through from the calibration's
`parameter_metadata` at their fixed values - every set is complete.

```{r posterior, eval = run_pest, message = FALSE, results = "hide"}
s3_calib  <- read_calib(ctrl = s3$cal_ctrl, sim_id = s3$cal_id)
post_sets <- pest_posterior_params(s3_calib)

b3_post  <- fresh_build("s3post", vars_bgc, use_bgc = TRUE)
aeme_ens <- run_aeme_ensemble(aeme = b3_post$aeme, model = MODEL,
                              path = b3_post$path, param_sets = post_sets,
                              parallel = TRUE, ncore = NCORE)
```

`run_aeme_ensemble()` gains a `param_sets` argument: given a list of `param`
dataframes (or a long data.frame with an `ensemble` id column) it runs those
instead of sampling from the bounds, and reuses the same parallel engine. The
result is the ordinary `AEME::output()$ens_*` structure.

`ensemble_summary()` collapses that structure once - looping the members
through `AEME::get_var()` a single time - into the per-date `mean`, `sd` and
quantile bands, keeping the per-member frame and the depth-aligned
observations alongside. `plot_ensemble()` takes either an `aeme` (it builds a
one-off summary internally, behaviour unchanged) or a summary directly, so one
extraction backs every plot, depth and interval, and `as.data.frame()` gives
the bands as a table:

```{r posterior-plot, eval = run_pest}
print(post_sets)
post_summary <- ensemble_summary(
  aeme_ens, model = MODEL,
  vars_sim = c("CHM_oxy", "NIT_amm", "PHS_frp"), depths = 20)

plot_ensemble(post_summary, var_sim = "CHM_oxy", depth = 20)
plot_ensemble(post_summary, var_sim = "NIT_amm", depth = 20, type = "line")
head(as.data.frame(post_summary))
```

`score_ensemble()` - coverage, ensemble-mean bias/RMSE, spread-skill ratio,
CRPS from the same object - is stubbed for a later release.

Two companions worth a look:

-   `pest_param_summary(s3_calib, param = ...)` - the prior-to-posterior
    variance-reduction table, i.e. which parameters the data actually
    constrained.
-   `get_sim_params(s3_calib, quantile_threshold = 0.1)` /
    `update_param(s3_calib, ...)` - a best-fit subset of every evaluated set,
    when you want a point estimate rather than the full ensemble.

### Resuming this step

The calibration is on disk: the results database and the kept PEST++ run
directory (`create_pest_control(keep_files = TRUE)`, the default). If a fresh
session only needs the ensemble, re-enter with

```r
s3_calib  <- read_calib(ctrl = s3$cal_ctrl, sim_id = s3$cal_id)
post_sets <- pest_posterior_params(s3_calib)
run_aeme_ensemble(aeme = fresh_build("s3post", vars_bgc, TRUE)$aeme,
                  model = MODEL, path = ..., param_sets = post_sets)
```

`inst/scripts/staged-calibration.R` checkpoints each stage to
`<ROOT>/sN.rds`, so `s3` comes back with `readRDS()`; if even that is gone,
`inst/scripts/staged-calibration-resume.R` rebuilds `s3$cal_ctrl` from the
results directory and reads `sim_id` out of the database. Run
`run_aeme_ensemble(parallel = FALSE)` if the installed package predates the
`make_temp_dir()` output-directory fix.

## Where to go next

-   `pest_adjust_weights()` rebalances observation groups after a first run so
    one variable does not dominate.
-   `validate_aeme()` scores the final parameter set over a held-out period;
    `get_calib_periods()` picks one.
-   Raise `morris_r`, `noptmax` and `ies_num_reals` well above the values used
    here before acting on any of it. These are sized to make a vignette build
    in minutes.
